---
title: "Transportation Chapter"
format: docx
execute:
  echo: false
  warning: false
  message: false
params:
  town: "Warren"
  town_code: 311
  state: "MA"
  county: "Worcester"
  analysis_year: 2023
  output_slug: "warren"
---
## NOTE: This is an automated report and should not be used exclusively as a final master plan chapter

# Roadway System Condition & Performance

## System Overview

```{r}
#| label: system-overview-road-inventory
#| results: "asis"

library(readr)
library(dplyr)
library(knitr)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

# Filter by numeric municipal code and counted in mileage
inv_town <- inv |>
  filter(City == params$town_code,
         Mile_Count == 1)

if (nrow(inv_town) == 0) {
  stop(paste("No rows found for town_code =", params$town_code))
}

summary_tbl <- inv_town |>
  mutate(
    roadway_miles = Length,
    lane_miles = Length * Num_Lanes
  ) |>
  summarise(
    roadway_miles = sum(roadway_miles, na.rm = TRUE),
    lane_miles = sum(lane_miles, na.rm = TRUE),
    segments = n()
  )

kable(summary_tbl, digits = 1,
      caption = paste0("Roadway miles and lane miles in ",
                       params$town,
                       " (MassDOT Road Inventory 2025)."))
```

```{r}
#| label: fig-miles-by-functional-class
#| fig-cap: "Roadway mileage by federal functional class (HPMS-counted segments)."
#| fig-height: 4.5
#| fig-width: 7

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

# Map functional class codes to labels (and combine 2 & 3)
ff_summary <- inv_town |>
  mutate(
    ff_label = case_when(
      F_F_Class == 1 ~ "Interstate",
      F_F_Class %in% c(2, 3) ~ "Principal Arterial",
      F_F_Class == 4 ~ "Minor Arterial",
      F_F_Class == 5 ~ "Major Collector",
      F_F_Class == 6 ~ "Minor Collector",
      F_F_Class == 7 ~ "Local",
      TRUE ~ "Other / Unknown"
    )
  ) |>
  group_by(ff_label) |>
  summarise(miles = sum(Length, na.rm = TRUE), .groups = "drop") |>
  # Order bars from most to least miles (cleaner in Word output)
  arrange(desc(miles))

ggplot(ff_summary, aes(x = reorder(ff_label, miles), y = miles)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Roadway miles",
    title = paste0(params$town, ": Roadway miles by federal functional class")
  )
```

```{r}
#| label: fig-miles-by-jurisdiction
#| fig-cap: "Roadway mileage by jurisdiction (HPMS-counted segments)."
#| fig-height: 4.5
#| fig-width: 7

library(readr)
library(dplyr)
library(ggplot2)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

jur_summary <- inv_town |>
  mutate(
    jur_label = case_when(
      Jurisdictn == "2" ~ "City/Town accepted road",
      Jurisdictn == "0" ~ "Unaccepted by city/town",
      Jurisdictn == "1" ~ "MassDOT",
      Jurisdictn == "3" ~ "DCR",
      Jurisdictn == "5" ~ "Massport",
      Jurisdictn == "6" ~ "State Park/Forest",
      Jurisdictn == "7" ~ "State Institutional",
      Jurisdictn == "8" ~ "Federal Park/Forest",
      Jurisdictn == "9" ~ "County Institutional",
      Jurisdictn == "B" ~ "State college/university",
      Jurisdictn == "C" ~ "US Air Force",
      Jurisdictn == "D" ~ "US Army Corps of Engineers",
      Jurisdictn == "E" ~ "Federal Institutional",
      Jurisdictn == "F" ~ "Other Federal",
      Jurisdictn == "G" ~ "Bureau of Indian Affairs",
      Jurisdictn == "H" ~ "Private",
      Jurisdictn == "I" ~ "US Army",
      Jurisdictn == "J" ~ "US Navy",
      TRUE ~ "Other/Unknown"
    )
  ) |>
  group_by(jur_label) |>
  summarise(miles = sum(Length, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(miles))

ggplot(jur_summary, aes(x = reorder(jur_label, miles), y = miles)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Roadway miles",
    title = paste0(params$town, ": Roadway miles by jurisdiction")
  )
```

## System Condition

```{r}
#| label: fig-miles-by-oci
#| fig-cap: "Miles of roadway by pavement condition category (OCI). Data reflects surveyed roadways only."
#| fig-height: 4.5
#| fig-width: 7

library(sf)
library(dplyr)
library(ggplot2)

pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
stopifnot(file.exists(pci_path))

pci <- st_read(pci_path, quiet = TRUE)

# Filter to town
pci_town <- pci |>
  filter(City == params$town_code)

if (nrow(pci_town) == 0) {
  stop(paste("No pavement records found for town_code =", params$town_code))
}

# Ensure projected CRS (important for length calculation)
if (st_is_longlat(pci_town)) {
  stop("Pavement layer is in geographic coordinates. Reproject to State Plane before exporting.")
}

# Calculate segment length in miles from geometry
pci_town <- pci_town |>
  mutate(
    length_miles = as.numeric(st_length(geometry)) / 1609.344
  )

# Categorize OCI
pci_summary <- pci_town |>
  mutate(
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    )
  ) |>
  group_by(oci_category) |>
  summarise(miles = sum(length_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(
    oci_category = factor(
      oci_category,
      levels = c("Very Poor", "Poor", "Fair", "Good", "Excellent")
    )
  )

ggplot(pci_summary, aes(x = oci_category, y = miles)) +
  geom_col() +
  geom_text(
    aes(label = round(miles, 1)),
    vjust = -0.3,
    size = 3.5
  ) +
  labs(
    x = NULL,
    y = "Miles",
    title = paste0(params$town, ": Pavement condition (OCI) distribution")
  ) +
  ylim(0, max(pci_summary$miles) * 1.15)

```

```{r}
#| label: map-oci
#| fig-cap: "Pavement condition (OCI) by roadway segment. Note: dataset reflects surveyed roadways only (primarily federal-aid eligible segments); many local roads are not included."
#| fig-height: 7.5
#| fig-width: 9
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(maptiles)

# paths
pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
bnd_path <- file.path("data_raw", "town_boundaries.geojson")  # <- change if needed

stopifnot(file.exists(pci_path))
stopifnot(file.exists(bnd_path))

# read + filter boundary
bnd <- st_read(bnd_path, quiet = TRUE) |>
  filter(TOWN_ID == params$town_code)

if (nrow(bnd) == 0) stop(paste("No boundary found for town_code =", params$town_code))

# read + filter pavement
pci <- st_read(pci_path, quiet = TRUE) |>
  filter(City == params$town_code)

if (nrow(pci) == 0) stop(paste("No pavement records found for town_code =", params$town_code))

# Categorize OCI
pci <- pci |>
  mutate(
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    ),
    oci_category = factor(
      oci_category,
      levels = c("Very Poor", "Poor", "Fair", "Good", "Excellent")
    )
  ) |>
  filter(!is.na(oci_category))

# Transform to lon/lat for web tiles
bnd_ll <- st_transform(bnd, 4326)
pci_ll <- st_transform(pci, 4326)

# Clip lines to the boundary so they don't bleed outside
pci_clip <- st_intersection(pci_ll, st_geometry(bnd_ll))


# Use the boundary bbox (not road bbox) for tiles/extents
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.1
ypad <- (bb["ymax"] - bb["ymin"]) * 0.1
bb_exp <- bb
bb_exp["xmin"] <- bb["xmin"] - xpad
bb_exp["xmax"] <- bb["xmax"] + xpad
bb_exp["ymin"] <- bb["ymin"] - ypad
bb_exp["ymax"] <- bb["ymax"] + ypad

# Higher zoom = less grainy (tweak 12–15 depending on town size)
tiles <- get_tiles(
  x = bb_exp,
  provider = "OpenStreetMap",
  zoom = 15,
  crop = FALSE
)

pal <- c(
  "Very Poor" = "#d73027",
  "Poor"      = "#fc8d59",
  "Fair"      = "#fee08b",
  "Good"      = "#d9ef8b",
  "Excellent" = "#1a9850"
)

# plot basemap
plot_tiles(tiles)

# boundary outline
plot(st_geometry(bnd_ll), add = TRUE, border = "black", lwd = 1)

# roads
for (cat in levels(pci_clip$oci_category)) {
  segs <- pci_clip[pci_clip$oci_category == cat, ]
  if (nrow(segs) > 0) plot(st_geometry(segs), add = TRUE, col = pal[[cat]], lwd = 4)
}

# Legend inside map: top-right corner of town bbox (inset)
bb <- st_bbox(bnd_ll)

x_pos <- bb["xmin"] - .012
y_pos <- bb["ymin"] + .02


legend(
  x = x_pos, y = y_pos,
  legend = names(pal),
  col = unname(pal),
  lwd = 3,
  bty = "o",
  bg = "white",
  box.lwd = 0.5,
  title = "Condition"
)

```


## System Maintenance and Financial Considerations

## Congestion

## Example Goals

# Community Travel Patterns and Accessibility

## Commute and Travel Modes

## Commute Times to Work

## Accessibility

## Walking, Bicycling & Transit Options

## Example Goals

# Safety & Speed Management

## Crash Locations and Severity

## Travel Speed

## Example Goals
