---
title: "Transportation Chapter"
format: 
  docx:
    fig-dpi: 300
execute:
  echo: false
  warning: false
  message: false
params:
  town: "Warren"
  town_code: 311
  state: "MA"
  county: "Worcester"
  analysis_year: 2023
  output_slug: "warren"
---

## NOTE: This is an automated report and should not be used exclusively as a final master plan chapter

```{r}
#| label: setup-plot-theme
#| include: false

library(ggplot2)

# Define a consistent color palette (planning-friendly)
tmp_palette <- c(
  primary = "#2C3E50",   # dark slate
  secondary = "#18BC9C", # muted teal
  accent = "#F39C12",    # soft orange
  light = "#ECF0F1"      # light gray
)

# Define a 3-color palette specifically for the historical trend charts
trend_palette <- c(
  "#D3D3D3", # Oldest period (light gray)
  "#18BC9C", # Middle period (teal)
  "#2C3E50"  # Most recent period (dark slate)
)

theme_tmp <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 13, face = "bold", color = tmp_palette["primary"]),
      plot.subtitle = element_text(size = 11, color = tmp_palette["primary"]),
      axis.title = element_text(size = 11, color = tmp_palette["primary"]),
      axis.text  = element_text(size = 11, color = tmp_palette["primary"]),
      legend.title = element_text(size = 11, color = tmp_palette["primary"]),
      legend.text  = element_text(size = 11),
      panel.grid.major = element_line(color = "#DADADA"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

theme_set(theme_tmp())
```

# Roadway System Condition & Performance

## System Overview

```{r}
#| label: system-overview-narrative
#| results: "asis"

library(readr)
library(dplyr)
library(epoxy)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

# Centerline miles
centerline_miles <- inv_town |>
  summarise(total = sum(Length, na.rm = TRUE)) |>
  pull(total)

# Lane miles
lane_miles <- inv_town |>
  mutate(lane_miles = Length * Num_Lanes) |>
  summarise(total = sum(lane_miles, na.rm = TRUE)) |>
  pull(total)

# Share of local functional class (lane miles basis)
local_share <- inv_town |>
  mutate(
    lane_miles = Length * Num_Lanes
  ) |>
  group_by(F_F_Class) |>
  summarise(miles = sum(lane_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(share = miles / sum(miles)) |>
  filter(F_F_Class == 7) |>
  pull(share)

# Share of MassDOT jurisdiction (lane miles basis)
massdot_share <- inv_town |>
  mutate(lane_miles = Length * Num_Lanes) |>
  group_by(Jurisdictn) |>
  summarise(miles = sum(lane_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(share = miles / sum(miles)) |>
  filter(Jurisdictn == 1) |>
  pull(share)

majority_owner_text <- if (local_share > 0.5) {
  "local governments maintain the majority of roadway mileage while MassDOT retains jurisdiction over higher-volume corridors"
} else {
  "state or other agencies maintain a larger-than-typical share of the roadway mileage, shifting some maintenance burden away from the municipality"
}

# Dynamic Logic: MassDOT presence
massdot_text <- if (length(massdot_share) > 0 && massdot_share > 0) {
  epoxy("MassDOT-owned corridors account for approximately {round(massdot_share * 100, 0)}% of total lane miles.")
} else {
  ""
}

# Output the text
epoxy("
The Town’s roadway network includes approximately {round(centerline_miles, 1)} centerline miles, representing roughly {round(lane_miles, 1)} total lane miles (MassDOT Road Inventory, 2025). Centerline mileage reflects the total length of roadway segments, while lane miles provide a more accurate measure of long-term maintenance responsibility and resurfacing needs.

Approximately {round(local_share * 100, 0)}% of total lane miles fall within the local functional classification, reflecting the Town’s predominantly residential character. {massdot_text} This structure is indicative of a system where {majority_owner_text}.

The distribution of roadway ownership and classification has important implications for maintenance responsibility, funding eligibility, and long-term capital planning.
")
```

```{r}
#| label: system-overview-road-inventory
#| results: "asis"

library(readr)
library(dplyr)
library(knitr)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

# Filter by numeric municipal code and counted in mileage
inv_town <- inv |>
  filter(City == params$town_code,
         Mile_Count == 1)

if (nrow(inv_town) == 0) {
  stop(paste("No rows found for town_code =", params$town_code))
}

summary_tbl <- inv_town |>
  mutate(
    roadway_miles = Length,
    lane_miles = Length * Num_Lanes
  ) |>
  summarise(
    roadway_miles = sum(roadway_miles, na.rm = TRUE),
    lane_miles = sum(lane_miles, na.rm = TRUE),
    segments = n()
  )

kable(summary_tbl, digits = 1,
      caption = paste0("Roadway miles and lane miles in ",
                       params$town,
                       " (MassDOT Road Inventory 2025)."))
```

```{r}
#| label: fig-miles-by-functional-class
#| fig-cap: "Centerline Roadway mileage by federal functional class (HPMS-counted segments)."
#| fig-height: 4.5
#| fig-width: 7

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

# Map functional class codes to labels (and combine 2 & 3)
ff_summary <- inv_town |>
  mutate(
    ff_label = case_when(
      F_F_Class == 1 ~ "Interstate",
      F_F_Class %in% c(2, 3) ~ "Principal Arterial",
      F_F_Class == 4 ~ "Minor Arterial",
      F_F_Class == 5 ~ "Major Collector",
      F_F_Class == 6 ~ "Minor Collector",
      F_F_Class == 7 ~ "Local",
      TRUE ~ "Other / Unknown"
    )
  ) |>
  group_by(ff_label) |>
  summarise(miles = sum(Length, na.rm = TRUE), .groups = "drop") |>
  # Order bars from most to least miles (cleaner in Word output)
  arrange(desc(miles))

ggplot(ff_summary, aes(x = reorder(ff_label, miles), y = miles)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Roadway miles",
    title = paste0(params$town, ": Roadway miles by federal functional class")
  )
```

```{r}
#| label: fig-miles-by-jurisdiction
#| fig-cap: "Centerline Roadway mileage by jurisdiction (HPMS-counted segments)."
#| fig-height: 3.5
#| fig-width: 7

library(readr)
library(dplyr)
library(ggplot2)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

jur_summary <- inv_town |>
  mutate(
    jur_label = case_when(
      Jurisdictn == "2" ~ "City/Town accepted road",
      Jurisdictn == "0" ~ "Unaccepted by city/town",
      Jurisdictn == "1" ~ "MassDOT",
      Jurisdictn == "3" ~ "DCR",
      Jurisdictn == "5" ~ "Massport",
      Jurisdictn == "6" ~ "State Park/Forest",
      Jurisdictn == "7" ~ "State Institutional",
      Jurisdictn == "8" ~ "Federal Park/Forest",
      Jurisdictn == "9" ~ "County Institutional",
      Jurisdictn == "B" ~ "State college/university",
      Jurisdictn == "C" ~ "US Air Force",
      Jurisdictn == "D" ~ "US Army Corps of Engineers",
      Jurisdictn == "E" ~ "Federal Institutional",
      Jurisdictn == "F" ~ "Other Federal",
      Jurisdictn == "G" ~ "Bureau of Indian Affairs",
      Jurisdictn == "H" ~ "Private",
      Jurisdictn == "I" ~ "US Army",
      Jurisdictn == "J" ~ "US Navy",
      TRUE ~ "Other/Unknown"
    )
  ) |>
  group_by(jur_label) |>
  summarise(miles = sum(Length, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(miles))

ggplot(jur_summary, aes(x = reorder(jur_label, miles), y = miles)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Roadway miles",
    title = paste0(params$town, ": Roadway miles by jurisdiction")
  )
```

## System Condition

```{r}
#| label: system-condition-narrative
#| results: "asis"

library(sf)
library(dplyr)
library(epoxy)

pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
pci <- st_read(pci_path, quiet = TRUE) |>
  filter(City == params$town_code)

if (nrow(pci) == 0) stop("No pavement records found for this town.")

if (st_is_longlat(pci)) {
  stop("Pavement layer must be projected (State Plane) before length calculation.")
}

# Calculate segment miles
pci <- pci |>
  mutate(
    length_miles = as.numeric(st_length(geometry)) / 1609.344,
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(oci_category))

total_miles <- sum(pci$length_miles, na.rm = TRUE)

summary_tbl <- pci |>
  group_by(oci_category) |>
  summarise(miles = sum(length_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(share = miles / total_miles)

# Pull category shares safely
get_share <- function(cat) {
  val <- summary_tbl$share[summary_tbl$oci_category == cat]
  ifelse(length(val) == 0, 0, val)
}

pct_excellent <- round(get_share("Excellent") * 100, 0)
pct_good      <- round(get_share("Good") * 100, 0)
pct_fair      <- round(get_share("Fair") * 100, 0)
pct_poor      <- round(get_share("Poor") * 100, 0)
pct_verypoor  <- round(get_share("Very Poor") * 100, 0)
pct_poor_combined <- pct_poor + pct_verypoor

good_fair_share <- pct_excellent + pct_good + pct_fair

condition_health_text <- if (good_fair_share >= 50) {
  "While a majority of roadways fall within excellent, good, or fair categories, even a modest share of poor-condition roads represents a significant long-term financial liability if not addressed proactively."
} else {
  "With a majority of the surveyed network falling into poor or very poor categories, the Town faces a substantial backlog of pavement needs that will require significant capital intervention."
}

# Output the text
epoxy("
Based on CMRPC pavement condition assessments, approximately {pct_excellent}% of surveyed roadway miles are rated in excellent condition, {pct_good}% in good condition, {pct_fair}% in fair condition, and {pct_poor_combined}% in poor or very poor condition. The dataset reflects approximately {round(total_miles, 1)} miles of surveyed roadway, primarily federal-aid eligible corridors.

{condition_health_text}

The spatial distribution of pavement conditions often reveals patterns — for example, older subdivision roads or rural collectors may exhibit more advanced deterioration, while recently resurfaced corridors perform better. Without a structured pavement management strategy, deferred maintenance can accelerate deterioration, increasing the eventual cost of reconstruction. Maintaining roadways in good or fair condition through timely resurfacing is typically more cost-effective than allowing conditions to decline to the point of full-depth reconstruction.
")
```

```{r}
#| label: fig-miles-by-oci
#| fig-cap: "Miles of roadway by pavement condition category (OCI). Data reflects surveyed roadways only."
#| fig-height: 4.5
#| fig-width: 7

library(sf)
library(dplyr)
library(ggplot2)

pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
stopifnot(file.exists(pci_path))

pci <- st_read(pci_path, quiet = TRUE)

# Filter to town
pci_town <- pci |>
  filter(City == params$town_code)

if (nrow(pci_town) == 0) {
  stop(paste("No pavement records found for town_code =", params$town_code))
}

# Ensure projected CRS (important for length calculation)
if (st_is_longlat(pci_town)) {
  stop("Pavement layer is in geographic coordinates. Reproject to State Plane before exporting.")
}

# Calculate segment length in miles from geometry
pci_town <- pci_town |>
  mutate(
    length_miles = as.numeric(st_length(geometry)) / 1609.344
  )

# Categorize OCI
pci_summary <- pci_town |>
  mutate(
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    )
  ) |>
  group_by(oci_category) |>
  summarise(miles = sum(length_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(
    oci_category = factor(
      oci_category,
      levels = c("Very Poor", "Poor", "Fair", "Good", "Excellent")
    )
  )

pal <- c(
  "Very Poor" = "#d73027",
  "Poor"      = "#fc8d59",
  "Fair"      = "#fee08b",
  "Good"      = "#d9ef8b",
  "Excellent" = "#1a9850"
)

tmp_palette <- c(
  primary = "#2C3E50",   # dark slate
  secondary = "#18BC9C", # muted teal
  accent = "#F39C12",    # soft orange
  light = "#ECF0F1"      # light gray
)

ggplot(pci_summary, aes(x = oci_category, y = miles, fill = oci_category)) +
  geom_col() +
  geom_text(
    aes(label = round(miles, 1)),
    vjust = -0.3,
    size = 3.5
  ) +
  labs(
    x = NULL,
    y = "Miles",
    title = paste0(params$town, ": Pavement condition (OCI) distribution")
  ) +
  ylim(0, max(pci_summary$miles) * 1.15) + 
  scale_fill_manual(values = pal, name = "OCI Category") +
   theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 13, face = "bold", color = tmp_palette["primary"]),
      plot.subtitle = element_text(size = 11, color = tmp_palette["primary"]),
      axis.title = element_text(size = 11, color = tmp_palette["primary"]),
      axis.text  = element_text(size = 11, color = tmp_palette["primary"]),
      legend.title = element_text(size = 11, color = tmp_palette["primary"]),
      legend.text  = element_text(size = 11),
      panel.grid.major = element_line(color = "#DADADA"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )

```

```{r}
#| label: map-oci
#| fig-cap: "Pavement condition (OCI) by roadway segment. Note: dataset reflects surveyed roadways only (primarily federal-aid eligible segments); many local roads are not included."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(ggplot2)
library(ggspatial)

# paths
pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
bnd_path <- file.path("data_raw", "town_boundaries.geojson")

stopifnot(file.exists(pci_path))
stopifnot(file.exists(bnd_path))

# read + filter boundary
bnd <- st_read(bnd_path, quiet = TRUE) |>
  filter(TOWN_ID == params$town_code)

if (nrow(bnd) == 0) stop(paste("No boundary found for town_code =", params$town_code))

# read + filter pavement
pci <- st_read(pci_path, quiet = TRUE) |>
  filter(City == params$town_code)

if (nrow(pci) == 0) stop(paste("No pavement records found for town_code =", params$town_code))

# Categorize OCI
pci <- pci |>
  mutate(
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    ),
    oci_category = factor(
      oci_category,
      levels = c("Very Poor", "Poor", "Fair", "Good", "Excellent")
    )
  ) |>
  filter(!is.na(oci_category))

# Transform to lon/lat for web tiles
bnd_ll <- st_transform(bnd, 4326)
pci_ll <- st_transform(pci, 4326)

# Clip lines to the boundary so they don't bleed outside
pci_clip <- st_intersection(pci_ll, st_geometry(bnd_ll))

# Define Palette
pal <- c(
  "Very Poor" = "#d73027",
  "Poor"      = "#fc8d59",
  "Fair"      = "#fee08b",
  "Good"      = "#d9ef8b",
  "Excellent" = "#1a9850"
)

# Set map limits with padding
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.05
ypad <- (bb["ymax"] - bb["ymin"]) * 0.05

# Build the map
ggplot() +
  # Add the basemap (uses OSM by default, similar to your maptiles setup)
  annotation_map_tile(type = "cartolight", zoom = 15, alpha = 0.8) +
  
  # Add the town boundary
  geom_sf(data = bnd_ll, fill = NA, color = "black", linewidth = 1) +
  
  # Add the clipped pavement segments
  geom_sf(data = pci_clip, aes(color = oci_category), linewidth = 1.5) +
  
  # Enforce the bounding box to keep the focus tight on the town
  coord_sf(
    xlim = c(bb["xmin"] - xpad, bb["xmax"] + xpad),
    ylim = c(bb["ymin"] - ypad, bb["ymax"] + ypad),
    expand = FALSE,
    crs = 4326
  ) +
  
  # Apply colors and legend title
  scale_color_manual(values = pal, drop = FALSE) +
  labs(color = "Condition") +
  
  # Styling for a clean, map-like appearance with an inset legend
  theme_void() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.15, 0.85), # Top-left inset (adjust coordinates as needed)
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    legend.margin = margin(6, 10, 6, 6),
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9)
  )
```

## System Maintenance and Financial Considerations

```{r}
#| label: data-prep
#| include: false

library(readr)
library(dplyr)
library(stringr)

# Consistent slug for filenames
town_slug <- params$town |>
  tolower() |>
  str_replace_all("\\s+", "_")

#### Clean Chapter 90
expected_filename <- paste0("chapter_90_", town_slug, ".csv")
ch90_path <- file.path("data_raw", expected_filename)

if (!file.exists(ch90_path)) {
  stop(paste0(
    "Chapter 90 file mismatch.\n",
    "Expected file: ", expected_filename, "\n",
    "Check that the Chapter 90 CSV matches params$town."
  ))
}

ch90_raw <- read_csv(ch90_path, show_col_types = FALSE)

ch90_clean <- ch90_raw |>
  mutate(fiscal_year = as.numeric(str_extract(Fiscal_Year, "\\d{4}"))) |>
  filter(!is.na(fiscal_year)) |>
  group_by(fiscal_year) |>
  summarise(
    chapter90_nominal = sum(Apportionment_Amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(fiscal_year)

#### Clean NHCCI (annual average from quarterly)
nhcci_path <- file.path("data_raw", "NHCCI_20260226.csv")
stopifnot(file.exists(nhcci_path))

nhcci_raw <- read_csv(nhcci_path, show_col_types = FALSE)

nhcci_clean <- nhcci_raw |>
  mutate(year = as.numeric(str_extract(quarter, "\\d{4}"))) |>
  filter(!is.na(year)) |>
  group_by(year) |>
  summarise(
    nhcci_index = mean(`NHCCI-Seasonally-Adjusted`, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(year)

#### Join + inflation adjust (base = earliest overlapping year)
ch90_overlap <- ch90_clean |>
  left_join(nhcci_clean, by = c("fiscal_year" = "year")) |>
  filter(!is.na(nhcci_index))

if (nrow(ch90_overlap) == 0) {
  stop("No overlapping years between Chapter 90 data and NHCCI index.")
}

base_year  <- min(ch90_overlap$fiscal_year, na.rm = TRUE)
base_index <- ch90_overlap |>
  filter(fiscal_year == base_year) |>
  pull(nhcci_index)

ch90_final <- ch90_overlap |>
  mutate(chapter90_real = chapter90_nominal * (base_index / nhcci_index))

```

```{r}
#| label: financial-narrative
#| results: "asis"

library(dplyr)
library(epoxy)

# Chapter 90 calculations

# Restrict to last 5 years available in adjusted dataset
latest_year <- max(ch90_final$fiscal_year, na.rm = TRUE)
five_year_start <- latest_year - 4

ch90_recent <- ch90_final |>
  filter(fiscal_year >= five_year_start)

avg_ch90_5yr <- mean(ch90_recent$chapter90_nominal, na.rm = TRUE)

# Trend direction (real)
real_change <- ch90_final$chapter90_real[ch90_final$fiscal_year == latest_year] -
               ch90_final$chapter90_real[ch90_final$fiscal_year == min(ch90_final$fiscal_year)]

# Dynamic Logic: Purchasing power trend
trend_text <- if (real_change > 50000) {
  "has seen a meaningful increase in real terms"
} else if (real_change >= -20000 && real_change <= 50000) {
  "has remained relatively stable in real terms"
} else {
  "has declined in real terms, meaning its purchasing power has dropped despite any nominal consistency"
}

# Specific example year
example_year <- 2016
example_text <- ""

if (example_year %in% ch90_final$fiscal_year) {
  ch90_2016_nominal <- ch90_final |>
    filter(fiscal_year == example_year) |>
    pull(chapter90_nominal)
  
  # Adjust 2016 to latest year dollars (reverse of base logic)
  nhcci_latest <- ch90_final |>
    filter(fiscal_year == latest_year) |>
    pull(nhcci_index)
  
  nhcci_2016 <- ch90_final |>
    filter(fiscal_year == example_year) |>
    pull(nhcci_index)
  
  ch90_2016_to_latest <- ch90_2016_nominal * (nhcci_latest / nhcci_2016)
  
  ch90_latest_nominal <- ch90_final |>
    filter(fiscal_year == latest_year) |>
    pull(chapter90_nominal)

  # Pre-build the example text if the year exists
  example_text <- epoxy("High levels of construction inflation mean that even though Chapter 90 funding has appeared relatively consistent in nominal terms, its purchasing power has shifted. For example, in FY{example_year}, Chapter 90 funding was ${format(round(ch90_2016_nominal, 0), big.mark = ',')}. If that amount had kept pace with highway construction inflation through FY{latest_year}, it would have been approximately ${format(round(ch90_2016_to_latest, 0), big.mark = ',')} in {latest_year} dollars, compared to actual FY{latest_year} funding of ${format(round(ch90_latest_nominal, 0), big.mark = ',')}. ")
}

# Output the text
epoxy("
The Town’s ability to maintain and improve its roadway network is closely tied to available funding. Over the past five years (FY{five_year_start}–FY{latest_year}), annual Chapter 90 allocations have averaged approximately ${format(round(avg_ch90_5yr, 0), big.mark = ',')}, supplemented by local public works expenditures.

When adjusted for inflation using the National Highway Construction Cost Index (NHCCI), Chapter 90 funding {trend_text} over the available analysis period. Figure X shows nominal and inflation-adjusted Chapter 90 funding, while Figure X presents the same comparison for General Fund Public Works spending.

{example_text}In many communities, the annual funding available for roadway maintenance is insufficient to address the full backlog of pavement needs. As a result, towns must balance routine maintenance, emergency repairs, and longer-term capital improvements within constrained budgets. Developing a multi-year pavement management plan that aligns available funding with prioritized roadway segments can help stabilize long-term costs and improve overall system performance.
")
```

```{r}
#| label: fig-ch90-nominal-vs-real
#| fig-cap: "Chapter 90 apportionment (nominal dollars and NHCCI-adjusted dollars; base year = earliest year with NHCCI data)."
#| fig-height: 4.5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)
library(tidyr)

plot_df <- ch90_final |>
  select(fiscal_year, chapter90_nominal, chapter90_real) |>
  pivot_longer(
    cols = c(chapter90_nominal, chapter90_real),
    names_to = "series",
    values_to = "amount"
  ) |>
  mutate(
    series = recode(
      series,
      chapter90_nominal = "Nominal ($)",
      chapter90_real    = paste0("NHCCI-adjusted (", base_year, " $)")
    )
  )

ggplot(plot_df, aes(x = fiscal_year, y = amount, group = series, linetype = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    labels = function(x) paste0("$", format(round(x / 1000, 0), big.mark = ","), "k")
  ) +
  labs(
    x = "Fiscal year",
    y = "dollars",
    title = paste0(params$town, ": Chapter 90 funding over time")
  )
```

```{r}
#| label: clean-and-adjust-general-fund
#| include: false

library(readxl)
library(dplyr)
library(stringr)
library(purrr)

gf_dir <- file.path("data_raw", "general_fund_expenditures")
stopifnot(dir.exists(gf_dir))

files <- list.files(
  gf_dir,
  pattern = "^GenFundExpenditures\\d{4}\\.xlsx$",
  full.names = TRUE
)

if (length(files) == 0) stop("No General Fund files found in data_raw/general_fund_expenditures/")

gf_all <- map_dfr(files, \(f) {
  readxl::read_excel(f, sheet = "Sheet1") |>
    transmute(
      municipality = Municipality,
      fiscal_year  = as.numeric(`Fiscal Year`),
      public_works_nominal = `Public Works`
    )
})

gf_town <- gf_all |>
  filter(str_to_upper(str_trim(municipality)) == str_to_upper(str_trim(params$town))) |>
  group_by(fiscal_year) |>
  summarise(
    public_works_nominal = sum(public_works_nominal, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(fiscal_year)

if (nrow(gf_town) == 0) {
  stop(paste0(
    "No General Fund rows found for town = '", params$town, "'. ",
    "Check Municipality spelling/casing in the General Fund files."
  ))
}

# Join to NHCCI
gf_joined <- gf_town |>
  left_join(nhcci_clean, by = c("fiscal_year" = "year"))

# Keep only overlapping years
gf_overlap <- gf_joined |>
  filter(!is.na(nhcci_index))

if (nrow(gf_overlap) == 0) {
  stop("No overlapping years between General Fund data and NHCCI index.")
}

# Base year = earliest overlapping year
gf_base_year <- min(gf_overlap$fiscal_year, na.rm = TRUE)
gf_base_index <- gf_overlap |>
  filter(fiscal_year == gf_base_year) |>
  pull(nhcci_index)

# Inflation adjust
gf_final <- gf_overlap |>
  mutate(
    public_works_real = public_works_nominal * (gf_base_index / nhcci_index)
  )
```

```{r}
#| label: fig-public-works-nominal-vs-real
#| fig-cap: "General Fund Public Works spending (nominal dollars and NHCCI-adjusted dollars; base year = earliest year with NHCCI data)."
#| fig-height: 4.5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)
library(tidyr)

plot_df_gf <- gf_final |>
  select(fiscal_year, public_works_nominal, public_works_real) |>
  pivot_longer(
    cols = c(public_works_nominal, public_works_real),
    names_to = "series",
    values_to = "amount"
  ) |>
  mutate(
    series = recode(
      series,
      public_works_nominal = "Nominal ($)",
      public_works_real    = paste0("NHCCI-adjusted (", gf_base_year, " $)")
    )
  )

ggplot(plot_df_gf, aes(x = fiscal_year, y = amount, group = series, linetype = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
  limits = c(100000, NA),
  labels = function(x) paste0("$", format(round(x / 1000, 0), big.mark = ","), "k")
) +
  labs(
    x = "Fiscal year",
    y = "dollars",
    title = paste0(params$town, ": General Fund Public Works spending over time")
  )
```

## Traffic Volume and Congestion

```{r}
#| label: replica-data-prep
#| include: false

library(sf)
library(dplyr)
library(stringr)

### File Paths

# Replica file (town-specific by municipal code)
replica_filename <- paste0("aadt_2024_", params$town_code, ".geojson")
replica_path <- file.path("data_raw", replica_filename)

if (!file.exists(replica_path)) {
  stop(paste0(
    "Replica AADT file mismatch.\n",
    "Expected file: ", replica_filename, "\n",
    "Check that the AADT GeoJSON matches params$town_code."
  ))
}

# Town boundary
bnd_path <- file.path("data_raw", "town_boundaries.geojson")
stopifnot(file.exists(bnd_path))

### Read Data

# Read boundary
bnd <- st_read(bnd_path, quiet = TRUE) |>
  filter(TOWN_ID == params$town_code)

if (nrow(bnd) == 0) {
  stop(paste("No boundary found for town_code =", params$town_code))
}

# Read Replica AADT data
rep <- st_read(replica_path, quiet = TRUE)

### Ensure Projected CRS (required for offsetting geometry)

if (st_is_longlat(rep)) {
  rep <- st_transform(rep, 26986)  # Massachusetts State Plane (meters)
}

if (st_is_longlat(bnd)) {
  bnd <- st_transform(bnd, 26986)
}

### Clip to Town Boundary and filter motorways out

rep_town <- st_intersection(rep, bnd) |>
  filter(highway != "motorway")

if (nrow(rep_town) == 0) {
  stop("No Replica AADT segments intersect town boundary.")
}

### Calculate Truck AADT and Share

rep_town <- rep_town |>
  mutate(
    aadt_single_unit = coalesce(aadt_single_unit, 0),
    aadt_combination = coalesce(aadt_combination, 0),
    truck_aadt = aadt_single_unit + aadt_combination,
    truck_share = ifelse(aadt > 0, truck_aadt / aadt, NA_real_)
  )


```

```{r}
#| label: tti-data-prep
#| include: false

library(sf)
library(dplyr)

### --- File Path (town-specific) ---

tti_filename <- paste0("qtr-hourly-speeds_2024_", params$town_code, ".geojson")
tti_path <- file.path("data_raw", tti_filename)

if (!file.exists(tti_path)) {
  stop(paste0(
    "Speed file mismatch.\n",
    "Expected file: ", tti_filename, "\n",
    "Check that the GeoJSON matches params$town_code."
  ))
}

#town boundary
bnd_path <- file.path("data_raw", "town_boundaries.geojson")  # <- change if needed

bnd <- st_read(bnd_path, quiet = TRUE) |>
  filter(TOWN_ID == params$town_code)

### --- Read Speeds Data ---

tti <- st_read(tti_path, quiet = TRUE)

# Ensure projected CRS (required for coordinate shifting)
if (st_is_longlat(tti)) {
  tti <- st_transform(tti, 26986)  # MA State Plane (meters)
}

### --- Identify Peak Period Columns ---

am_cols <- paste0("wkdy_", c(
  "0700","0715","0730","0745",
  "0800","0815","0830","0845"
))

pm_cols <- paste0("wkdy_", c(
  "1600","1615","1630","1645",
  "1700","1715","1730","1745"
))

### --- Compute Peak Speeds ---

tti <- tti |>
  mutate(
    am_peak_speed = rowMeans(across(all_of(am_cols)), na.rm = TRUE),
    pm_peak_speed = rowMeans(across(all_of(pm_cols)), na.rm = TRUE)
  )

### --- Compute Travel Time Index (TTI) ---

tti <- tti |>
  mutate(
    am_tti = free_flow_speed_mph / am_peak_speed,
    pm_tti = free_flow_speed_mph / pm_peak_speed
  )

### --- Offset Directional Links Using Heading Only (no extra sign split) ---

dist_m <- 15  # increase if needed (e.g., 12)

offset_geom_by_heading <- function(geom, heading_deg, dist_m) {
  if (is.na(heading_deg)) heading_deg <- 0

  theta <- heading_deg * pi / 180

  # Direction vector (0=N, 90=E)
  dx <- sin(theta)
  dy <- cos(theta)

  # Perpendicular (right of travel direction)
  ox <-  dy * dist_m
  oy <- -dx * dist_m

  shift_coords <- function(mat) {
    mat[, 1] <- mat[, 1] + ox
    mat[, 2] <- mat[, 2] + oy
    mat
  }

  if (inherits(geom, "LINESTRING")) {
    return(st_linestring(shift_coords(geom)))
  }

  if (inherits(geom, "MULTILINESTRING")) {
    parts <- lapply(geom, shift_coords)
    return(st_multilinestring(parts))
  }

  geom
}

geom_list <- st_geometry(tti)
hdg <- tti$heading

tti_offset <- tti
st_geometry(tti_offset) <- st_sfc(
  lapply(seq_along(geom_list), function(i) {
    offset_geom_by_heading(geom_list[[i]], hdg[i], dist_m)
  }),
  crs = st_crs(tti)
)
```

```{r}
#| label: townwide-tti-data
#| include: false

library(dplyr)
library(tidyr)

tti_town <- tti |>
  filter(highway != "motorway")

qtr_cols <- grep("^wkdy_", names(tti_town), value = TRUE)

tti_long <- tti_town |>
  select(length, free_flow_speed_mph, all_of(qtr_cols)) |>
  pivot_longer(
    cols = all_of(qtr_cols),
    names_to = "time",
    values_to = "observed_speed"
  ) |>
  filter(!is.na(observed_speed), observed_speed > 0) |>
  mutate(
    link_tti = free_flow_speed_mph / observed_speed
  )

townwide_tti <- tti_long |>
  group_by(time) |>
  summarise(
    town_tti = weighted.mean(link_tti, w = length, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(time)
```

```{r}
#| label: txt-traffic-overview
#| results: "asis"
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(glue)
library(epoxy)

# Basic network coverage counts (excluding motorways, consistent with maps)
n_aadt_links <- rep_town |>
  filter(highway != "motorway") |>
  nrow()

n_speed_links <- tti |>
  filter(highway != "motorway") |>
  nrow()

epoxy("
Traffic volumes and congestion patterns provide insight into how the roadway network functions during a typical day, and where delays may affect daily travel. In {params$town}, traffic activity is concentrated on a limited set of higher-order corridors, while many local streets primarily serve neighborhood access. The analyses below summarize estimated traffic volumes (AADT) and truck activity from Replica, and evaluate congestion using the Travel Time Index (TTI), which compares peak-period travel conditions to free-flow conditions.
")
```

```{r}
#| label: txt-aadt-volumes
#| results: "asis"
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(epoxy)

aadt_df <- rep_town |>
  filter(highway != "motorway") |>
  st_drop_geometry() |>
  mutate(
    truck_aadt = coalesce(aadt_single_unit, 0) + coalesce(aadt_combination, 0),
    truck_share = ifelse(aadt > 0, truck_aadt / aadt, NA)
  )

# Simple townwide stats
aadt_median <- median(aadt_df$aadt, na.rm = TRUE)
aadt_p90 <- as.numeric(quantile(aadt_df$aadt, 0.90, na.rm = TRUE))
truck_share_med <- median(aadt_df$truck_share, na.rm = TRUE)

# Top corridor by AADT (by street_name, excluding missing)
top_corr <- aadt_df |>
  filter(!is.na(street_name), street_name != "") |>
  group_by(street_name) |>
  summarise(aadt = max(aadt, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(aadt)) |>
  slice(1)

top_name <- if (nrow(top_corr) == 1) top_corr$street_name else "key corridors"
top_aadt <- if (nrow(top_corr) == 1) top_corr$aadt else NA_real_

fmt_int <- function(x) format(round(x, 0), big.mark = ",")

# Dynamic Logic: Top corridor mention
top_corridor_text <- if (!is.na(top_aadt)) {
  epoxy("The highest volumes are observed along {top_name} (approximately {fmt_int(top_aadt)} vehicles per day on the busiest segments). ")
} else {
  ""
}

# Output the text
epoxy("
Estimated traffic volumes vary widely across the network. The median segment in {params$town} carries approximately {fmt_int(aadt_median)} vehicles per day, while higher-volume corridors reach roughly {fmt_int(aadt_p90)} vehicles per day (90th percentile). 

{top_corridor_text}Truck activity is generally concentrated on the same corridors; the typical segment has an estimated truck share of about {round(truck_share_med * 100, 1)}% of daily traffic. Figure X shows total AADT (all vehicles) and figure X shows the estimated proportion of traffic made up of trucks.
")
```

```{r}
#| label: map-aadt-total
#| fig-cap: "Estimated Annual Average Daily Traffic (AADT), all vehicles (Replica, 2024). Highways (motorway segments) are excluded to emphasize local roadway conditions."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(ggplot2)
library(ggspatial)

# Transform to lon/lat for tiles + plotting
bnd_ll <- st_transform(bnd, 4326)

# Filter out motorways (keep motorway_link) and transform
rep_ll <- rep_town |>
  filter(highway != "motorway") |>
  st_transform(4326)

# Manual breaks 
rep_ll <- rep_ll |>
  mutate(
    aadt_cat = cut(
      aadt,
      breaks = c(-Inf, 500, 1000, 2000, 3000, Inf),
      labels = c("<500", "500 – 1k", "1 – 2k", "2 – 3k", "3k+")
    )
  )

# Graduated color ramp (light -> dark)
col_map <- c(
  "<500"     = "#F6BDC0",
  "500 – 1k" = "#F1959B",
  "1 – 2k"   = "#F07470",
  "2 – 3k"   = "#EA4C46",
  "3k+"      = "#DC1C13"
)

# Line widths (thin -> thick)
lwd_map <- c(
  "<500"     = 0.8, 
  "500 – 1k" = 1.2, 
  "1 – 2k"   = 1.8, 
  "2 – 3k"   = 2.5, 
  "3k+"      = 3.5
)

# Set map limits with padding
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.05
ypad <- (bb["ymax"] - bb["ymin"]) * 0.05

ggplot() +
  # Using cartolight for a clean, muted background
  annotation_map_tile(type = "cartolight", zoom = 15, alpha = 0.9) +
  
  # Add the AADT segments, mapping both color and line thickness
  geom_sf(data = rep_ll, aes(color = aadt_cat, linewidth = aadt_cat)) +

  # Add the town boundary
  geom_sf(data = bnd_ll, fill = NA, color = "black", linewidth = 1) +
    
  # Enforce the bounding box to keep the focus tight on the town
  coord_sf(
    xlim = c(bb["xmin"] - xpad, bb["xmax"] + xpad),
    ylim = c(bb["ymin"] - ypad, bb["ymax"] + ypad),
    expand = FALSE,
    crs = 4326
  ) +
  
  # Apply scales
  scale_color_manual(values = col_map, drop = FALSE) +
  scale_linewidth_manual(values = lwd_map, drop = FALSE) +
  
  # Combine color and linewidth into a single legend title
  labs(color = "AADT", linewidth = "AADT") +
  
  # Map styling
  theme_void() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.10, 0.82), # Top-left inset
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    legend.margin = margin(6, 12, 6, 8),
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10)
  )
```

```{r}
#| label: map-aadt-truck-share
#| fig-cap: "Truck share of AADT (single-unit + combination trucks as a percentage of total AADT; Replica, 2024). Highways (motorway segments) are excluded to emphasize local roadway conditions."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(ggplot2)
library(ggspatial)

# Transform boundary to lon/lat for tiles + plotting
bnd_ll <- st_transform(bnd, 4326)

# Filter out motorways (keep motorway_link) and transform
rep_ll <- rep_town |>
  filter(highway != "motorway") |>
  st_transform(4326)

# Define the exact categories we want everywhere
cat_levels <- c("<1%", "1–2%", "2–5%", "5–10%", "10%+")

# Calculate truck percentage and categorize (TRIPLE-LOCK STEP 1: Strict Factor)
rep_ll <- rep_ll |>
  mutate(
    truck_pct = truck_share * 100,
    truck_cat = cut(
      truck_pct,
      breaks = c(-Inf, 1, 2, 5, 10, Inf),
      labels = cat_levels
    ),
    # Force it to be a factor with all levels explicitly stated
    truck_cat = factor(truck_cat, levels = cat_levels)
  ) |>
  filter(!is.na(truck_cat))

# Color ramp: green (low) -> red (high)
col_map <- c(
  "<1%"   = "#1a9850",
  "1–2%"  = "#91cf60",
  "2–5%"  = "#fee08b",
  "5–10%" = "#fc8d59",
  "10%+"  = "#d73027"
)

# Line widths (adjusted for ggplot2: thin -> thick)
lwd_map <- c(
  "<1%"   = 0.8,
  "1–2%"  = 1.2,
  "2–5%"  = 1.8,
  "5–10%" = 2.5,
  "10%+"  = 3.5
)

# Set map limits with padding
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.05
ypad <- (bb["ymax"] - bb["ymin"]) * 0.05

ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 15, alpha = 0.9) +
  geom_sf(data = rep_ll, aes(color = truck_cat, linewidth = truck_cat), show.legend = TRUE) +
  geom_sf(data = bnd_ll, fill = NA, color = "black", linewidth = 1) +
  coord_sf(
    xlim = c(bb["xmin"] - xpad, bb["xmax"] + xpad),
    ylim = c(bb["ymin"] - ypad, bb["ymax"] + ypad),
    expand = FALSE,
    crs = 4326
  ) +
  
  # TRIPLE-LOCK STEP 2 & 3: drop = FALSE AND explicit breaks/limits
  scale_color_manual(
    values = col_map, 
    drop = FALSE, 
    breaks = cat_levels, 
    limits = cat_levels
  ) +
  scale_linewidth_manual(
    values = lwd_map, 
    drop = FALSE, 
    breaks = cat_levels, 
    limits = cat_levels
  ) +
  
  labs(color = "Truck Share", linewidth = "Truck Share") +
  theme_void() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.12, 0.82), 
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    legend.margin = margin(6, 12, 6, 8),
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10)
  )
```

```{r}
#| label: txt-tti-congestion
#| results: "asis"
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(glue)

# Peak-period townwide TTI from the chart dataset
townwide_tti2 <- townwide_tti |>
  mutate(t = sub("^wkdy_", "", time))

am_times <- c("0700","0715","0730","0745","0800","0815","0830","0845")
pm_times <- c("1600","1615","1630","1645","1700","1715","1730","1745")

am_tti_avg <- townwide_tti2 |>
  filter(t %in% am_times) |>
  summarise(v = mean(town_tti, na.rm = TRUE)) |>
  pull(v)

pm_tti_avg <- townwide_tti2 |>
  filter(t %in% pm_times) |>
  summarise(v = mean(town_tti, na.rm = TRUE)) |>
  pull(v)

tti_max <- max(townwide_tti2$town_tti, na.rm = TRUE)

qtr_cols <- grep("^wkdy_", names(tti), value = TRUE)
am_cols <- paste0("wkdy_", am_times)

am_link <- tti |>
  filter(highway != "motorway") |>
  st_drop_geometry() |>
  transmute(
    free_flow_speed_mph,
    am_peak_speed = rowMeans(across(all_of(am_cols)), na.rm = TRUE)
  ) |>
  filter(!is.na(am_peak_speed), am_peak_speed > 0, !is.na(free_flow_speed_mph), free_flow_speed_mph > 0) |>
  mutate(am_tti = free_flow_speed_mph / am_peak_speed)

# Here is the variable that was missing!
share_am_over_115 <- mean(am_link$am_tti > 1.15, na.rm = TRUE)

# Dynamic Logic: Congestion severity scale
delay_severity_text <- if (share_am_over_115 < 0.05) {
  "experiences very limited recurring delay"
} else if (share_am_over_115 < 0.15) {
  "experiences moderate localized delay, typical of suburban peak periods"
} else {
  "experiences significant and widespread recurring delay across the network"
}

epoxy("
Congestion is summarized using the Travel Time Index (TTI), where a value of 1.00 indicates free-flow conditions and a value of 1.20 indicates travel times approximately 20% longer than free flow. Based on Replica weekday speeds, the townwide average TTI is approximately {round(am_tti_avg, 2)} during the AM peak (7:00–9:00 AM) and {round(pm_tti_avg, 2)} during the PM peak (4:00–6:00 PM), with the highest quarter-hour townwide value reaching about {round(tti_max, 2)}.

In general, the network {delay_severity_text}; during the AM peak, about {round(share_am_over_115 * 100, 0)}% of analyzed segments exceed a TTI of 1.15 (roughly 15% longer travel time than free flow). Map X highlights where peak-period delay is most likely to occur, while Figure X shows how townwide congestion changes by time of day.
")
```

```{r}
#| label: map-tti-am
#| fig-cap: "AM Peak Travel Time Index (TTI), weekdays 7:00–9:00 AM. TTI compares peak travel time to free-flow travel time (TTI = 1.2 indicates ~20% longer travel time than free flow)."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(ggplot2)
library(ggspatial)

# Transform to lon/lat for tiles + plotting
bnd_ll <- st_transform(bnd, 4326)

tti_ll <- tti_offset |>
  filter(highway != "motorway") |>
  st_transform(4326)

# Define categories
cat_levels <- c("≤1.05", "1.05–1.15", "1.15–1.30", "1.30–1.50", "1.50+")

# Categorize and lock as factor
tti_ll <- tti_ll |>
  mutate(
    am_tti_cat = cut(
      am_tti,
      breaks = c(-Inf, 1.05, 1.15, 1.30, 1.50, Inf),
      labels = cat_levels
    ),
    am_tti_cat = factor(am_tti_cat, levels = cat_levels)
  ) |>
  filter(!is.na(am_tti_cat))

# Color ramp: green (low delay) -> red (high delay)
col_map <- c(
  "≤1.05"     = "#1a9850",
  "1.05–1.15" = "#91cf60",
  "1.15–1.30" = "#fee08b",
  "1.30–1.50" = "#fc8d59",
  "1.50+"     = "#d73027"
)

# Set map limits with padding
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.05
ypad <- (bb["ymax"] - bb["ymin"]) * 0.05

ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 15, alpha = 0.9) +
  
  # TTI segments with the show.legend = TRUE fix applied!
  geom_sf(data = tti_ll, aes(color = am_tti_cat), linewidth = 0.5, show.legend = TRUE) +
  
  # Town boundary
  geom_sf(data = bnd_ll, fill = NA, color = "black", linewidth = 1) +
  
  coord_sf(
    xlim = c(bb["xmin"] - xpad, bb["xmax"] + xpad),
    ylim = c(bb["ymin"] - ypad, bb["ymax"] + ypad),
    expand = FALSE,
    crs = 4326
  ) +
  
  # Apply scale with drop = FALSE
  scale_color_manual(values = col_map, drop = FALSE) +
  # Override the legend aesthetics to make the lines thicker
  guides(color = guide_legend(override.aes = list(linewidth = 1))) +
  
  labs(color = "AM Peak TTI") +
  
  theme_void() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.12, 0.82), 
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    legend.margin = margin(6, 12, 6, 8),
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10)
  )
```

```{r}
#| label: plot-townwide-tti
#| fig-cap: "Townwide weekday Travel Time Index (TTI) by time of day. TTI = 1.2 indicates travel times approximately 20% longer than free-flow conditions."
#| fig-height: 4.5
#| fig-width: 9
#| fig-dpi: 300

library(ggplot2)
library(dplyr)
library(lubridate)

townwide_tti <- townwide_tti |>
  mutate(
    time_clean = sub("wkdy_", "", time),
    time_posix = as.POSIXct(time_clean, format = "%H%M")
  ) |>
  arrange(time_posix)

# Define peak windows
am_start <- as.POSIXct("07:00", format = "%H:%M")
am_end   <- as.POSIXct("09:00", format = "%H:%M")
pm_start <- as.POSIXct("16:00", format = "%H:%M")
pm_end   <- as.POSIXct("18:00", format = "%H:%M")

# Choose a label height slightly below max
label_y <- max(townwide_tti$town_tti, na.rm = TRUE) * 0.98

ggplot(townwide_tti, aes(x = time_posix, y = town_tti)) +
  # Shaded peak periods
  annotate("rect", xmin = am_start, xmax = am_end, ymin = -Inf, ymax = Inf, alpha = 0.12) +
  annotate("rect", xmin = pm_start, xmax = pm_end, ymin = -Inf, ymax = Inf, alpha = 0.12) +
  
  # Peak labels
  annotate("text",
           x = am_start + (am_end - am_start)/2,
           y = label_y,
           label = "AM Peak\n(7–9 AM)",
           vjust = 1,
           size = 3.5) +
  annotate("text",
           x = pm_start + (pm_end - pm_start)/2,
           y = label_y,
           label = "PM Peak\n(4–6 PM)",
           vjust = 1,
           size = 3.5) +
  
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  
  scale_x_datetime(
    date_labels = "%H:%M",
    date_breaks = "3 hours"
  ) +
  scale_y_continuous(
    limits = c(0.9, max(townwide_tti$town_tti, na.rm = TRUE) * 1.05)
  ) +
  
  labs(
    x = "Time of Day",
    y = "Travel Time Index (TTI)",
    title = paste0(params$town, ": Townwide Travel Time Index by Time of Day")
  )
```

## Roadway System Condition & Performanc Goals

{{< pagebreak >}}

# Community Travel Patterns and Accessibility

## Commute and Travel Modes

```{r}
#| label: data-prep-acs
#| include: false

library(tidycensus)
library(dplyr)
library(stringr)
library(purrr)

# 1. Define your variables
vars_commute <- c(
  "Drive Alone"      = "B08301_003",
  "Carpool"          = "B08301_004",
  "Public Transit"   = "B08301_010",
  "Taxicab"          = "B08301_016",
  "Motorcycle"       = "B08301_017",
  "Bicycle"          = "B08301_018",
  "Walked"           = "B08301_019",
  "Other"            = "B08301_020",
  "Worked from Home" = "B08301_021"
)

vars_vehicle <- c(
  "No vehicle" = "B08201_002",
  "1 vehicle"  = "B08201_003",
  "2 vehicles" = "B08201_004",
  "3 vehicles" = "B08201_005",
  "4+ vehicles"= "B08201_006"
)

vars_travel_time <- c(
  "< 5 min"   = "B08303_002",
  "5-9 min"   = "B08303_003",
  "10-14 min" = "B08303_004",
  "15-19 min" = "B08303_005",
  "20-24 min" = "B08303_006",
  "25-29 min" = "B08303_007",
  "30-34 min" = "B08303_008",
  "35-39 min" = "B08303_009",
  "40-44 min" = "B08303_010",
  "45-59 min" = "B08303_011",
  "60-89 min" = "B08303_012",
  "90+ min"   = "B08303_013"
)

all_vars <- c(vars_commute, vars_vehicle, vars_travel_time)

# 2. Dynamically find the latest available ACS 5-year data
current_year <- as.integer(format(Sys.Date(), "%Y"))
latest_acs_year <- NA

# Ping the API with a tiny, fast state-level request, stepping back from the current year
for (yr in current_year:2015) {
  test_call <- try(
    suppressMessages(get_acs(
      geography = "state", state = "MA", variables = "B08301_001", year = yr, survey = "acs5", quiet = TRUE
    )),
    silent = TRUE
  )
  
  if (!inherits(test_call, "try-error")) {
    latest_acs_year <- yr
    break
  }
}

if (is.na(latest_acs_year)) stop("Could not connect to Census API to find the latest year.")

# Calculate the 3 non-overlapping years based on the anchor
target_years <- c(
  latest_acs_year, 
  latest_acs_year - 5, 
  latest_acs_year - 10
)

# Print a helpful message to the console during rendering
message(paste("\n---> Using ACS 5-Year Data for:", paste(target_years, collapse = ", "), "<---\n"))

# 3. Pull data mapped across the 3 years
acs_raw <- map_dfr(target_years, function(yr) {
  get_acs(
    geography = "county subdivision",
    state = params$state,
    county = params$county,
    variables = all_vars,
    year = yr,
    survey = "acs5",
    quiet = TRUE
  ) |>
    mutate(year = yr) # Tag the data with the year it was pulled
})

# 4. Filter to the specific town
acs_town <- acs_raw |>
  mutate(
    town_clean = str_extract(NAME, "^[^,]+") |> 
                 str_remove_all("(?i)\\s+(town|city)$") |> 
                 str_trim()
  ) |>
  filter(str_to_upper(town_clean) == str_to_upper(params$town))

if(nrow(acs_town) == 0) stop(paste("No ACS data found for town:", params$town))

# 5. Split out datasets
df_commute <- acs_town |> filter(variable %in% names(vars_commute))
df_vehicle <- acs_town |> filter(variable %in% names(vars_vehicle))
df_time    <- acs_town |> filter(variable %in% names(vars_travel_time))

```

```{r}
#| label: txt-commute-modes
#| results: "asis"
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(epoxy)

# Identify years for trend comparison
latest_acs_year <- max(df_commute$year, na.rm = TRUE)
oldest_acs_year <- min(df_commute$year, na.rm = TRUE)

# Calculate total commuters for the latest year
commute_totals <- df_commute |> 
  group_by(year) |> 
  summarise(total = sum(estimate, na.rm = TRUE), .groups = "drop")

# Helper function to get % share
get_mode_pct <- function(yr, mode_name) {
  val <- df_commute |> filter(year == yr, variable == mode_name) |> pull(estimate) |> sum(na.rm = TRUE)
  tot <- commute_totals |> filter(year == yr) |> pull(total)
  if(length(val) == 0 || tot == 0) return(0)
  return(val / tot)
}

# Extract key statistics
drive_alone_pct <- round(get_mode_pct(latest_acs_year, "Drive Alone") * 100, 1)
carpool_pct <- round(get_mode_pct(latest_acs_year, "Carpool") * 100, 1)
wfh_latest <- get_mode_pct(latest_acs_year, "Worked from Home")
wfh_oldest <- get_mode_pct(oldest_acs_year, "Worked from Home")

wfh_pct_display <- round(wfh_latest * 100, 1)

# Dynamic Logic: Remote Work Trend
wfh_trend_text <- if (wfh_latest > (wfh_oldest + 0.03)) {
  "remote work saw a notable increase over the past decade, likely accelerated by the COVID-19 pandemic, and has maintained a higher long-term share"
} else {
  "while remote work fluctuated during the COVID-19 pandemic, recent trends indicate minimal long-term change from previous baselines"
}

# Dynamic Logic: Vehicle Availability
veh_tot <- df_vehicle |> filter(year == latest_acs_year) |> pull(estimate) |> sum(na.rm = TRUE)
veh_0 <- df_vehicle |> filter(year == latest_acs_year, variable == "No vehicle") |> pull(estimate) |> sum(na.rm = TRUE)
veh_0_pct <- if(veh_tot > 0) round((veh_0 / veh_tot) * 100, 1) else 0

epoxy("
According to recent American Community Survey (ACS) data, the majority of employed residents commute to work by driving alone, accounting for approximately {drive_alone_pct}% of all work trips. Carpooling represents a smaller share at {carpool_pct}%, while walking, bicycling, and public transportation account for comparatively modest proportions. Over the past decade, commute mode share has remained relatively stable, with limited shifts away from single-occupancy vehicle travel. 

This pattern is consistent with many small and mid-sized Central Massachusetts communities, where regional job dispersion and limited transit availability constrain non-automobile travel options. Furthermore, {wfh_trend_text}, currently representing {wfh_pct_display}% of the workforce. Household vehicle availability strongly mirrors this car-dependent travel pattern; only {veh_0_pct}% of households report having zero vehicles available.
")
```


```{r}
#| label: fig-commute-mode
#| fig-cap: "Primary means of transportation to work over time (ACS 5-Year Estimates)."
#| fig-height: 5.5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)

# Clean up categories and calculate shares PER YEAR
df_commute_plot <- df_commute |>
  mutate(
    plot_category = case_when(
      variable %in% c("Taxicab", "Motorcycle", "Other") ~ "Other",
      TRUE ~ variable
    )
  ) |>
  group_by(year, plot_category) |>
  summarise(estimate = sum(estimate, na.rm = TRUE), .groups = "drop_last") |>
  mutate(
    share = estimate / sum(estimate),
    # Create clean labels like "2019-2023"
    year_label = paste0(year - 4, "–", year) 
  ) |>
  ungroup()

# Sort the y-axis categories based on the MOST RECENT year's data
recent_order <- df_commute_plot |>
  filter(year == max(year)) |>
  arrange(share) |>
  pull(plot_category)

df_commute_plot <- df_commute_plot |>
  mutate(plot_category = factor(plot_category, levels = recent_order))

ggplot(df_commute_plot, aes(x = plot_category, y = share, fill = as.factor(year_label))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = trend_palette) + 
  labs(
    x = NULL,
    y = "Share of Commuters",
    fill = "5-Year Period",
    title = paste0(params$town, ": Means of Transportation to Work"),
    subtitle = "Comparing non-overlapping ACS 5-Year Estimates"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```


```{r}
#| label: fig-vehicle-availability
#| fig-cap: "Household vehicle availability over time (ACS 5-Year Estimates)."
#| fig-height: 4.5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)

# 1. Define strict factor levels for logical ordering
veh_levels <- c("No vehicle", "1 vehicle", "2 vehicles", "3 vehicles", "4+ vehicles")

# 2. Calculate shares per year
df_vehicle_plot <- df_vehicle |>
  group_by(year, variable) |>
  summarise(estimate = sum(estimate, na.rm = TRUE), .groups = "drop_last") |>
  mutate(
    share = estimate / sum(estimate),
    year_label = paste0(year - 4, "–", year),
    variable = factor(variable, levels = veh_levels)
  ) |>
  ungroup()

# 3. Plot
ggplot(df_vehicle_plot, aes(x = variable, y = share, fill = as.factor(year_label))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = trend_palette) + 
  labs(
    x = NULL,
    y = "Share of Households",
    fill = "5-Year Period",
    title = paste0(params$town, ": Vehicle Availability"),
    subtitle = "Comparing non-overlapping ACS 5-Year Estimates"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```
```{r}
#| label: data-prep-lodes
#| include: false
#| cache: true

library(tidyverse)
library(readr)
library(stringr)

# 1. Helper Functions to read directly from LEHD servers
read_lodes_od <- function(state, year, main = TRUE, base = "https://lehd.ces.census.gov/data/lodes/LODES8/") {
  state <- tolower(state)
  type_str <- if(main) "main" else "aux"
  file_url <- paste0(base, state, "/od/", state, "_od_", type_str, "_JT00_", year, ".csv.gz")
  
  read_csv(
    file_url, 
    col_types = cols(h_geocode = col_character(), w_geocode = col_character()), 
    show_col_types = FALSE,
    progress = FALSE
  )
}

read_crosswalk <- function(state, base = "https://lehd.ces.census.gov/data/lodes/LODES8/") {
  state <- tolower(state)
  file_url <- paste0(base, state, "/", state, "_xwalk.csv.gz")
  
  read_csv(
    file_url, 
    col_types = cols(.default = col_character()), 
    col_select = c("tabblk2020", "ctycsubname"), 
    show_col_types = FALSE,
    progress = FALSE
  )
}

# 2. Fetch Data (Using 2021 as it is widely available across all states)
lodes_state <- tolower(params$state)
lodes_year <- 2023 

od_main <- read_lodes_od(lodes_state, lodes_year, main = TRUE)
crosswalk <- read_crosswalk(lodes_state)

# 3. Clean Crosswalk Town Names to perfectly match params$town
crosswalk_clean <- crosswalk |>
  mutate(
    town_clean = str_remove(ctycsubname, "(?i)\\s+(town|city)\\s+\\(.*$") |> 
                 str_trim()
  )

# 4. Join home and work geographies
od_joined <- od_main |>
  left_join(crosswalk_clean, by = c("h_geocode" = "tabblk2020")) |>
  rename(home_town = town_clean) |>
  left_join(crosswalk_clean, by = c("w_geocode" = "tabblk2020")) |>
  rename(work_town = town_clean)

# 5. Filter for residents of the parameter town and find top 10 destinations
od_top_destinations <- od_joined |>
  filter(str_to_upper(home_town) == str_to_upper(params$town)) |>
  group_by(work_town) |>
  summarise(total_commuters = sum(S000, na.rm = TRUE), .groups = "drop") |>
  # Handle NA values (e.g., out-of-state commutes missing from a single-state crosswalk)
  mutate(work_town = ifelse(is.na(work_town), "Out of State / Other", work_town)) |>
  group_by(work_town) |>
  summarise(total_commuters = sum(total_commuters), .groups = "drop") |>
  slice_max(order_by = total_commuters, n = 10)
```

## Commute Times to Work

```{r}
#| label: txt-commute-times
#| results: "asis"
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(epoxy)

# Calculate commute times over 30 mins for the most recent year
latest_time_year <- max(df_time$year, na.rm = TRUE)
time_tot <- df_time |> filter(year == latest_time_year) |> pull(estimate) |> sum(na.rm = TRUE)

long_commute_vars <- c("30-34 min", "35-39 min", "40-44 min", "45-59 min", "60-89 min", "90+ min")
time_30plus <- df_time |> 
  filter(year == latest_time_year, variable %in% long_commute_vars) |> 
  pull(estimate) |> 
  sum(na.rm = TRUE)

commute_30plus_pct <- if(time_tot > 0) round((time_30plus / time_tot) * 100, 1) else 0

# Extract Top 3 LODES Destinations
top_3_dests <- od_top_destinations |> slice_head(n = 3) |> pull(work_town)

# Format the list nicely (e.g., "Worcester, Auburn, and Millbury")
dest_string <- if(length(top_3_dests) > 2) {
  paste0(paste(top_3_dests[-length(top_3_dests)], collapse = ", "), ", and ", top_3_dests[length(top_3_dests)])
} else {
  paste(top_3_dests, collapse = " and ")
}

epoxy("
A substantial share of workers face longer daily travel durations, with approximately {commute_30plus_pct}% commuting 30 minutes or more to reach their primary place of employment. LODES inflow/outflow analysis indicates that a significant portion of employed residents travel outside the Town for work, particularly to regional employment centers such as {dest_string}. 

Longer commute times are often linked to these regional employment patterns rather than local congestion alone. In many communities, extended commute durations reflect the sheer distance traveled to major job hubs rather than delay within the local roadway network. Understanding this geographic distribution of job access helps inform land use planning, economic development strategy, and future transportation investments.
")
```


```{r}
#| label: fig-travel-time
#| fig-cap: "Commute travel time to work over time (ACS 5-Year Estimates)."
#| fig-height: 5.5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)

# 1. Group the detailed categories into logical 15-minute bins
df_time_plot <- df_time |>
  mutate(
    plot_category = case_when(
      variable %in% c("< 5 min", "5-9 min", "10-14 min") ~ "< 15 min",
      variable %in% c("15-19 min", "20-24 min", "25-29 min") ~ "15–29 min",
      variable %in% c("30-34 min", "35-39 min", "40-44 min") ~ "30–44 min",
      variable == "45-59 min" ~ "45–59 min",
      variable %in% c("60-89 min", "90+ min") ~ "60+ min",
      TRUE ~ "Other"
    )
  ) |>
  # 2. Summarize the new bins per year
  group_by(year, plot_category) |>
  summarise(estimate = sum(estimate, na.rm = TRUE), .groups = "drop_last") |>
  mutate(
    share = estimate / sum(estimate),
    year_label = paste0(year - 4, "–", year)
  ) |>
  ungroup()

# 3. Define strict chronological factor levels for the new bins
time_levels <- c("< 15 min", "15–29 min", "30–44 min", "45–59 min", "60+ min")

df_time_plot <- df_time_plot |>
  mutate(
    # Reverse the factor levels here so the shortest time is at the TOP of the flipped chart
    plot_category = factor(plot_category, levels = rev(time_levels)) 
  )

# 4. Plot
ggplot(df_time_plot, aes(x = plot_category, y = share, fill = as.factor(year_label))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = trend_palette) + 
  labs(
    x = NULL,
    y = "Share of Commuters",
    fill = "5-Year Period",
    title = paste0(params$town, ": Commute Travel Time"),
    subtitle = "Comparing non-overlapping ACS 5-Year Estimates"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```


```{r}
#| label: fig-commute-destinations
#| fig-cap: "Top 10 work destinations for residents (LODES Origin-Destination Data)."
#| fig-height: 5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)

# Order factors from largest to smallest for plotting
plot_lodes <- od_top_destinations |>
  mutate(work_town = factor(work_town, levels = rev(work_town)))

ggplot(plot_lodes, aes(x = work_town, y = total_commuters)) +
  # Using the primary color from your global theme
  geom_col(fill = tmp_palette["primary"], width = 0.7) +
  # Add labels to the ends of the bars for quick readability
  geom_text(aes(label = scales::comma(total_commuters)), 
            hjust = -0.2, size = 3.5, color = tmp_palette["primary"]) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma_format(), 
    expand = expansion(mult = c(0, 0.15)) # Give labels breathing room on the right
  ) +
  labs(
    x = "Work Destination",
    y = "Number of Commuters",
    title = paste0(params$town, ": Top 10 Work Destinations"),
    subtitle = paste0("LODES Origin-Destination Data (", lodes_year, ")")
  ) +
  theme(
    panel.grid.major.y = element_blank() # Remove horizontal grid lines for a cleaner bar chart
  )
```

## Accessibility

## Walking, Bicycling & Transit Options

## Example Goals

# Safety & Speed Management

## Crash Locations and Severity

## Travel Speed

## Example Goals
