---
title: "Transportation Chapter"
format: 
  docx:
    fig-dpi: 300
execute:
  echo: false
  warning: false
  message: false
params:
  town: "Warren"
  town_code: 311
  state: "MA"
  county: "Worcester"
  analysis_year: 2023
  output_slug: "warren"
---
## NOTE: This is an automated report and should not be used exclusively as a final master plan chapter

```{r}
#| label: setup-plot-theme
#| include: false

library(ggplot2)

# Define a consistent color palette (planning-friendly)
tmp_palette <- c(
  primary = "#2C3E50",   # dark slate
  secondary = "#18BC9C", # muted teal
  accent = "#F39C12",    # soft orange
  light = "#ECF0F1"      # light gray
)

theme_tmp <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 13, face = "bold", color = tmp_palette["primary"]),
      plot.subtitle = element_text(size = 11, color = tmp_palette["primary"]),
      axis.title = element_text(size = 11, color = tmp_palette["primary"]),
      axis.text  = element_text(size = 11, color = tmp_palette["primary"]),
      legend.title = element_text(size = 11, color = tmp_palette["primary"]),
      legend.text  = element_text(size = 11),
      panel.grid.major = element_line(color = "#DADADA"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

theme_set(theme_tmp())
```


# Roadway System Condition & Performance

## System Overview
```{r}
#| label: system-overview-narrative
#| results: "asis"

library(readr)
library(dplyr)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

# Centerline miles
centerline_miles <- inv_town |>
  summarise(total = sum(Length, na.rm = TRUE)) |>
  pull(total)

# Lane miles
lane_miles <- inv_town |>
  mutate(lane_miles = Length * Num_Lanes) |>
  summarise(total = sum(lane_miles, na.rm = TRUE)) |>
  pull(total)

# Share of local functional class (lane miles basis)
local_share <- inv_town |>
  mutate(
    lane_miles = Length * Num_Lanes
  ) |>
  group_by(F_F_Class) |>
  summarise(miles = sum(lane_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(share = miles / sum(miles)) |>
  filter(F_F_Class == 7) |>
  pull(share)

# Share of MassDOT jurisdiction (lane miles basis)
massdot_share <- inv_town |>
  mutate(lane_miles = Length * Num_Lanes) |>
  group_by(Jurisdictn) |>
  summarise(miles = sum(lane_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(share = miles / sum(miles)) |>
  filter(Jurisdictn == 1) |>
  pull(share)

cat(paste0(
  "The Town’s roadway network includes approximately ",
  round(centerline_miles, 1),
  " centerline miles, representing roughly ",
  round(lane_miles, 1),
  " total lane miles (MassDOT Road Inventory, 2025). ",
  "Centerline mileage reflects the total length of roadway segments, ",
  "while lane miles provide a more accurate measure of long-term maintenance responsibility and resurfacing needs. ",
  "Approximately ",
  round(local_share * 100, 0),
  "% of total lane miles fall within the local functional classification, ",
  "reflecting the Town’s predominantly residential character. ",
  ifelse(length(massdot_share) == 0,
         "",
         paste0("MassDOT-owned corridors account for approximately ",
                round(massdot_share * 100, 0),
                "% of total lane miles. ")
  ),
  "This structure is typical of many Central Massachusetts communities, ",
  "where local governments maintain the majority of roadway mileage while MassDOT retains jurisdiction over higher-volume corridors. ",
  "The distribution of roadway ownership and classification has important implications for maintenance responsibility, funding eligibility, and long-term capital planning."
))
```

```{r}
#| label: system-overview-road-inventory
#| results: "asis"

library(readr)
library(dplyr)
library(knitr)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

# Filter by numeric municipal code and counted in mileage
inv_town <- inv |>
  filter(City == params$town_code,
         Mile_Count == 1)

if (nrow(inv_town) == 0) {
  stop(paste("No rows found for town_code =", params$town_code))
}

summary_tbl <- inv_town |>
  mutate(
    roadway_miles = Length,
    lane_miles = Length * Num_Lanes
  ) |>
  summarise(
    roadway_miles = sum(roadway_miles, na.rm = TRUE),
    lane_miles = sum(lane_miles, na.rm = TRUE),
    segments = n()
  )

kable(summary_tbl, digits = 1,
      caption = paste0("Roadway miles and lane miles in ",
                       params$town,
                       " (MassDOT Road Inventory 2025)."))
```

```{r}
#| label: fig-miles-by-functional-class
#| fig-cap: "Centerline Roadway mileage by federal functional class (HPMS-counted segments)."
#| fig-height: 4.5
#| fig-width: 7

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

# Map functional class codes to labels (and combine 2 & 3)
ff_summary <- inv_town |>
  mutate(
    ff_label = case_when(
      F_F_Class == 1 ~ "Interstate",
      F_F_Class %in% c(2, 3) ~ "Principal Arterial",
      F_F_Class == 4 ~ "Minor Arterial",
      F_F_Class == 5 ~ "Major Collector",
      F_F_Class == 6 ~ "Minor Collector",
      F_F_Class == 7 ~ "Local",
      TRUE ~ "Other / Unknown"
    )
  ) |>
  group_by(ff_label) |>
  summarise(miles = sum(Length, na.rm = TRUE), .groups = "drop") |>
  # Order bars from most to least miles (cleaner in Word output)
  arrange(desc(miles))

ggplot(ff_summary, aes(x = reorder(ff_label, miles), y = miles)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Roadway miles",
    title = paste0(params$town, ": Roadway miles by federal functional class")
  )
```

```{r}
#| label: fig-miles-by-jurisdiction
#| fig-cap: "Centerline Roadway mileage by jurisdiction (HPMS-counted segments)."
#| fig-height: 3.5
#| fig-width: 7

library(readr)
library(dplyr)
library(ggplot2)

inv_path <- file.path("data_raw", "Road_Inventory_2025.csv")
stopifnot(file.exists(inv_path))

inv <- read_csv(inv_path, show_col_types = FALSE)

inv_town <- inv |>
  filter(
    City == params$town_code,
    Mile_Count == 1
  )

jur_summary <- inv_town |>
  mutate(
    jur_label = case_when(
      Jurisdictn == "2" ~ "City/Town accepted road",
      Jurisdictn == "0" ~ "Unaccepted by city/town",
      Jurisdictn == "1" ~ "MassDOT",
      Jurisdictn == "3" ~ "DCR",
      Jurisdictn == "5" ~ "Massport",
      Jurisdictn == "6" ~ "State Park/Forest",
      Jurisdictn == "7" ~ "State Institutional",
      Jurisdictn == "8" ~ "Federal Park/Forest",
      Jurisdictn == "9" ~ "County Institutional",
      Jurisdictn == "B" ~ "State college/university",
      Jurisdictn == "C" ~ "US Air Force",
      Jurisdictn == "D" ~ "US Army Corps of Engineers",
      Jurisdictn == "E" ~ "Federal Institutional",
      Jurisdictn == "F" ~ "Other Federal",
      Jurisdictn == "G" ~ "Bureau of Indian Affairs",
      Jurisdictn == "H" ~ "Private",
      Jurisdictn == "I" ~ "US Army",
      Jurisdictn == "J" ~ "US Navy",
      TRUE ~ "Other/Unknown"
    )
  ) |>
  group_by(jur_label) |>
  summarise(miles = sum(Length, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(miles))

ggplot(jur_summary, aes(x = reorder(jur_label, miles), y = miles)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Roadway miles",
    title = paste0(params$town, ": Roadway miles by jurisdiction")
  )
```

## System Condition
```{r}
#| label: system-condition-narrative
#| results: "asis"

library(sf)
library(dplyr)

pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
pci <- st_read(pci_path, quiet = TRUE) |>
  filter(City == params$town_code)

if (nrow(pci) == 0) stop("No pavement records found for this town.")

if (st_is_longlat(pci)) {
  stop("Pavement layer must be projected (State Plane) before length calculation.")
}

# Calculate segment miles
pci <- pci |>
  mutate(
    length_miles = as.numeric(st_length(geometry)) / 1609.344,
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(oci_category))

total_miles <- sum(pci$length_miles, na.rm = TRUE)

summary_tbl <- pci |>
  group_by(oci_category) |>
  summarise(miles = sum(length_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(share = miles / total_miles)

# Pull category shares safely
get_share <- function(cat) {
  val <- summary_tbl$share[summary_tbl$oci_category == cat]
  ifelse(length(val) == 0, 0, val)
}

pct_excellent <- round(get_share("Excellent") * 100, 0)
pct_good      <- round(get_share("Good") * 100, 0)
pct_fair      <- round(get_share("Fair") * 100, 0)
pct_poor      <- round(get_share("Poor") * 100, 0)
pct_verypoor  <- round(get_share("Very Poor") * 100, 0)
pct_poor_combined <- pct_poor + pct_verypoor

cat(paste0(
  "Based on CMRPC pavement condition assessments, approximately ",
  pct_excellent, "% of surveyed roadway miles are rated in excellent condition, ",
  pct_good, "% in good condition, ",
  pct_fair, "% in fair condition, and ",
  pct_poor_combined, "% in poor or very poor condition. ",
  "The dataset reflects approximately ",
  round(total_miles, 1),
  " miles of surveyed roadway, primarily federal-aid eligible corridors. ",
  "While a majority of roadways fall within excellent, good, or fair categories, ",
  "even a modest share of poor-condition roads represents a significant long-term financial liability if not addressed proactively.\n\n",
  "The spatial distribution of pavement conditions often reveals patterns — for example, older subdivision roads or rural collectors may exhibit more advanced deterioration, ",
  "while recently resurfaced corridors perform better. Without a structured pavement management strategy, deferred maintenance can accelerate deterioration, increasing the eventual cost of reconstruction. ",
  "Maintaining roadways in good or fair condition through timely resurfacing is typically more cost-effective than allowing conditions to decline to the point of full-depth reconstruction."
))
```

```{r}
#| label: fig-miles-by-oci
#| fig-cap: "Miles of roadway by pavement condition category (OCI). Data reflects surveyed roadways only."
#| fig-height: 4.5
#| fig-width: 7

library(sf)
library(dplyr)
library(ggplot2)

pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
stopifnot(file.exists(pci_path))

pci <- st_read(pci_path, quiet = TRUE)

# Filter to town
pci_town <- pci |>
  filter(City == params$town_code)

if (nrow(pci_town) == 0) {
  stop(paste("No pavement records found for town_code =", params$town_code))
}

# Ensure projected CRS (important for length calculation)
if (st_is_longlat(pci_town)) {
  stop("Pavement layer is in geographic coordinates. Reproject to State Plane before exporting.")
}

# Calculate segment length in miles from geometry
pci_town <- pci_town |>
  mutate(
    length_miles = as.numeric(st_length(geometry)) / 1609.344
  )

# Categorize OCI
pci_summary <- pci_town |>
  mutate(
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    )
  ) |>
  group_by(oci_category) |>
  summarise(miles = sum(length_miles, na.rm = TRUE), .groups = "drop") |>
  mutate(
    oci_category = factor(
      oci_category,
      levels = c("Very Poor", "Poor", "Fair", "Good", "Excellent")
    )
  )

pal <- c(
  "Very Poor" = "#d73027",
  "Poor"      = "#fc8d59",
  "Fair"      = "#fee08b",
  "Good"      = "#d9ef8b",
  "Excellent" = "#1a9850"
)

tmp_palette <- c(
  primary = "#2C3E50",   # dark slate
  secondary = "#18BC9C", # muted teal
  accent = "#F39C12",    # soft orange
  light = "#ECF0F1"      # light gray
)

ggplot(pci_summary, aes(x = oci_category, y = miles, fill = oci_category)) +
  geom_col() +
  geom_text(
    aes(label = round(miles, 1)),
    vjust = -0.3,
    size = 3.5
  ) +
  labs(
    x = NULL,
    y = "Miles",
    title = paste0(params$town, ": Pavement condition (OCI) distribution")
  ) +
  ylim(0, max(pci_summary$miles) * 1.15) + 
  scale_fill_manual(values = pal, name = "OCI Category") +
   theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 13, face = "bold", color = tmp_palette["primary"]),
      plot.subtitle = element_text(size = 11, color = tmp_palette["primary"]),
      axis.title = element_text(size = 11, color = tmp_palette["primary"]),
      axis.text  = element_text(size = 11, color = tmp_palette["primary"]),
      legend.title = element_text(size = 11, color = tmp_palette["primary"]),
      legend.text  = element_text(size = 11),
      panel.grid.major = element_line(color = "#DADADA"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )

```

```{r}
#| label: map-oci
#| fig-cap: "Pavement condition (OCI) by roadway segment. Note: dataset reflects surveyed roadways only (primarily federal-aid eligible segments); many local roads are not included."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(maptiles)

# paths
pci_path <- file.path("data_raw", "pavement_cartegraph.geojson")
bnd_path <- file.path("data_raw", "town_boundaries.geojson")  # <- change if needed

stopifnot(file.exists(pci_path))
stopifnot(file.exists(bnd_path))

# read + filter boundary
bnd <- st_read(bnd_path, quiet = TRUE) |>
  filter(TOWN_ID == params$town_code)

if (nrow(bnd) == 0) stop(paste("No boundary found for town_code =", params$town_code))

# read + filter pavement
pci <- st_read(pci_path, quiet = TRUE) |>
  filter(City == params$town_code)

if (nrow(pci) == 0) stop(paste("No pavement records found for town_code =", params$town_code))

# Categorize OCI
pci <- pci |>
  mutate(
    oci_category = case_when(
      Survey_OCI >= 0  & Survey_OCI < 24 ~ "Very Poor",
      Survey_OCI >= 24 & Survey_OCI < 47 ~ "Poor",
      Survey_OCI >= 47 & Survey_OCI < 67 ~ "Fair",
      Survey_OCI >= 67 & Survey_OCI < 87 ~ "Good",
      Survey_OCI >= 87                   ~ "Excellent",
      TRUE ~ NA_character_
    ),
    oci_category = factor(
      oci_category,
      levels = c("Very Poor", "Poor", "Fair", "Good", "Excellent")
    )
  ) |>
  filter(!is.na(oci_category))

# Transform to lon/lat for web tiles
bnd_ll <- st_transform(bnd, 4326)
pci_ll <- st_transform(pci, 4326)

# Clip lines to the boundary so they don't bleed outside
pci_clip <- st_intersection(pci_ll, st_geometry(bnd_ll))


# Use the boundary bbox (not road bbox) for tiles/extents
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.1
ypad <- (bb["ymax"] - bb["ymin"]) * 0.1
bb_exp <- bb
bb_exp["xmin"] <- bb["xmin"] - xpad
bb_exp["xmax"] <- bb["xmax"] + xpad
bb_exp["ymin"] <- bb["ymin"] - ypad
bb_exp["ymax"] <- bb["ymax"] + ypad

# Higher zoom = less grainy (tweak 12–15 depending on town size)
tiles <- get_tiles(
  x = bb_exp,
  provider = "OpenStreetMap",
  zoom = 15,
  crop = FALSE
)

pal <- c(
  "Very Poor" = "#d73027",
  "Poor"      = "#fc8d59",
  "Fair"      = "#fee08b",
  "Good"      = "#d9ef8b",
  "Excellent" = "#1a9850"
)

# plot basemap
plot_tiles(tiles)

# boundary outline
plot(st_geometry(bnd_ll), add = TRUE, border = "black", lwd = 1)

# roads
for (cat in levels(pci_clip$oci_category)) {
  segs <- pci_clip[pci_clip$oci_category == cat, ]
  if (nrow(segs) > 0) plot(st_geometry(segs), add = TRUE, col = pal[[cat]], lwd = 4)
}

# Legend inside map: top-right corner of town bbox (inset)
bb <- st_bbox(bnd_ll)

x_pos <- bb["xmin"] - .012
y_pos <- bb["ymin"] + .02


legend(
  x = x_pos, y = y_pos,
  legend = names(pal),
  col = unname(pal),
  lwd = 3,
  bty = "o",
  bg = "white",
  box.lwd = 0.5,
  title = "Condition"
)

```

## System Maintenance and Financial Considerations
```{r}
#| label: data-prep
#| include: false

library(readr)
library(dplyr)
library(stringr)

# Consistent slug for filenames
town_slug <- params$town |>
  tolower() |>
  str_replace_all("\\s+", "_")

#### Clean Chapter 90
expected_filename <- paste0("chapter_90_", town_slug, ".csv")
ch90_path <- file.path("data_raw", expected_filename)

if (!file.exists(ch90_path)) {
  stop(paste0(
    "Chapter 90 file mismatch.\n",
    "Expected file: ", expected_filename, "\n",
    "Check that the Chapter 90 CSV matches params$town."
  ))
}

ch90_raw <- read_csv(ch90_path, show_col_types = FALSE)

ch90_clean <- ch90_raw |>
  mutate(fiscal_year = as.numeric(str_extract(Fiscal_Year, "\\d{4}"))) |>
  filter(!is.na(fiscal_year)) |>
  group_by(fiscal_year) |>
  summarise(
    chapter90_nominal = sum(Apportionment_Amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(fiscal_year)

#### Clean NHCCI (annual average from quarterly)
nhcci_path <- file.path("data_raw", "NHCCI_20260226.csv")
stopifnot(file.exists(nhcci_path))

nhcci_raw <- read_csv(nhcci_path, show_col_types = FALSE)

nhcci_clean <- nhcci_raw |>
  mutate(year = as.numeric(str_extract(quarter, "\\d{4}"))) |>
  filter(!is.na(year)) |>
  group_by(year) |>
  summarise(
    nhcci_index = mean(`NHCCI-Seasonally-Adjusted`, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(year)

#### Join + inflation adjust (base = earliest overlapping year)
ch90_overlap <- ch90_clean |>
  left_join(nhcci_clean, by = c("fiscal_year" = "year")) |>
  filter(!is.na(nhcci_index))

if (nrow(ch90_overlap) == 0) {
  stop("No overlapping years between Chapter 90 data and NHCCI index.")
}

base_year  <- min(ch90_overlap$fiscal_year, na.rm = TRUE)
base_index <- ch90_overlap |>
  filter(fiscal_year == base_year) |>
  pull(nhcci_index)

ch90_final <- ch90_overlap |>
  mutate(chapter90_real = chapter90_nominal * (base_index / nhcci_index))

```

```{r}
#| label: financial-narrative
#| results: "asis"

library(dplyr)

# Chapter 90 calculations

# Restrict to last 5 years available in adjusted dataset
latest_year <- max(ch90_final$fiscal_year, na.rm = TRUE)
five_year_start <- latest_year - 4

ch90_recent <- ch90_final |>
  filter(fiscal_year >= five_year_start)

avg_ch90_5yr <- mean(ch90_recent$chapter90_nominal, na.rm = TRUE)

# Trend direction (real)
real_change <- ch90_final$chapter90_real[ch90_final$fiscal_year == latest_year] -
               ch90_final$chapter90_real[ch90_final$fiscal_year == min(ch90_final$fiscal_year)]

trend_text <- ifelse(real_change >= 0,
                     "has remained relatively stable or increased modestly",
                     "has declined in real terms")

# Specific example year
example_year <- 2016

if (example_year %in% ch90_final$fiscal_year) {
  ch90_2016_nominal <- ch90_final |>
    filter(fiscal_year == example_year) |>
    pull(chapter90_nominal)
  
  # Adjust 2016 to latest year dollars (reverse of base logic)
  nhcci_latest <- ch90_final |>
    filter(fiscal_year == latest_year) |>
    pull(nhcci_index)
  
  nhcci_2016 <- ch90_final |>
    filter(fiscal_year == example_year) |>
    pull(nhcci_index)
  
  ch90_2016_to_latest <- ch90_2016_nominal * (nhcci_latest / nhcci_2016)
  
} else {
  ch90_2016_nominal <- NA
  ch90_2016_to_latest <- NA
}

ch90_latest_nominal <- ch90_final |>
  filter(fiscal_year == latest_year) |>
  pull(chapter90_nominal)

cat(paste0(
  "The Town’s ability to maintain and improve its roadway network is closely tied to available funding. ",
  "Over the past five years (FY", five_year_start, "–FY", latest_year, "), annual Chapter 90 allocations have averaged approximately $",
  format(round(avg_ch90_5yr, 0), big.mark = ","), 
  ", supplemented by local public works expenditures. ",
  "When adjusted for inflation using the National Highway Construction Cost Index (NHCCI), Chapter 90 funding ",
  trend_text, 
  " over the available analysis period. ",
  "Figure X shows nominal and inflation-adjusted Chapter 90 funding, while Figure X presents the same comparison for General Fund Public Works spending.\n\n",
  ifelse(!is.na(ch90_2016_nominal),
         paste0(
           "High levels of construction inflation mean that even though Chapter 90 funding has appeared relatively consistent in nominal terms, its purchasing power has shifted. ",
           "For example, in FY", example_year, ", Chapter 90 funding was $",
           format(round(ch90_2016_nominal, 0), big.mark = ","), 
           ". If that amount had kept pace with highway construction inflation through FY", latest_year,
           ", it would have been approximately $",
           format(round(ch90_2016_to_latest, 0), big.mark = ","), 
           " in ", latest_year, " dollars, compared to actual FY", latest_year,
           " funding of $", format(round(ch90_latest_nominal, 0), big.mark = ","), 
           ". "
         ),
         ""
  ),
  "In many communities, the annual funding available for roadway maintenance is insufficient to address the full backlog of pavement needs. ",
  "As a result, towns must balance routine maintenance, emergency repairs, and longer-term capital improvements within constrained budgets. ",
  "Developing a multi-year pavement management plan that aligns available funding with prioritized roadway segments can help stabilize long-term costs and improve overall system performance."
))
```

```{r}
#| label: fig-ch90-nominal-vs-real
#| fig-cap: "Chapter 90 apportionment (nominal dollars and NHCCI-adjusted dollars; base year = earliest year with NHCCI data)."
#| fig-height: 4.5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)
library(tidyr)

plot_df <- ch90_final |>
  select(fiscal_year, chapter90_nominal, chapter90_real) |>
  pivot_longer(
    cols = c(chapter90_nominal, chapter90_real),
    names_to = "series",
    values_to = "amount"
  ) |>
  mutate(
    series = recode(
      series,
      chapter90_nominal = "Nominal ($)",
      chapter90_real    = paste0("NHCCI-adjusted (", base_year, " $)")
    )
  )

ggplot(plot_df, aes(x = fiscal_year, y = amount, group = series, linetype = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    labels = function(x) paste0("$", format(round(x / 1000, 0), big.mark = ","), "k")
  ) +
  labs(
    x = "Fiscal year",
    y = "dollars",
    title = paste0(params$town, ": Chapter 90 funding over time")
  )
```

```{r}
#| label: clean-and-adjust-general-fund
#| include: false

library(readxl)
library(dplyr)
library(stringr)
library(purrr)

gf_dir <- file.path("data_raw", "general_fund_expenditures")
stopifnot(dir.exists(gf_dir))

files <- list.files(
  gf_dir,
  pattern = "^GenFundExpenditures\\d{4}\\.xlsx$",
  full.names = TRUE
)

if (length(files) == 0) stop("No General Fund files found in data_raw/general_fund_expenditures/")

gf_all <- map_dfr(files, \(f) {
  readxl::read_excel(f, sheet = "Sheet1") |>
    transmute(
      municipality = Municipality,
      fiscal_year  = as.numeric(`Fiscal Year`),
      public_works_nominal = `Public Works`
    )
})

gf_town <- gf_all |>
  filter(str_to_upper(str_trim(municipality)) == str_to_upper(str_trim(params$town))) |>
  group_by(fiscal_year) |>
  summarise(
    public_works_nominal = sum(public_works_nominal, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(fiscal_year)

if (nrow(gf_town) == 0) {
  stop(paste0(
    "No General Fund rows found for town = '", params$town, "'. ",
    "Check Municipality spelling/casing in the General Fund files."
  ))
}

# Join to NHCCI
gf_joined <- gf_town |>
  left_join(nhcci_clean, by = c("fiscal_year" = "year"))

# Keep only overlapping years
gf_overlap <- gf_joined |>
  filter(!is.na(nhcci_index))

if (nrow(gf_overlap) == 0) {
  stop("No overlapping years between General Fund data and NHCCI index.")
}

# Base year = earliest overlapping year
gf_base_year <- min(gf_overlap$fiscal_year, na.rm = TRUE)
gf_base_index <- gf_overlap |>
  filter(fiscal_year == gf_base_year) |>
  pull(nhcci_index)

# Inflation adjust
gf_final <- gf_overlap |>
  mutate(
    public_works_real = public_works_nominal * (gf_base_index / nhcci_index)
  )
```

```{r}
#| label: fig-public-works-nominal-vs-real
#| fig-cap: "General Fund Public Works spending (nominal dollars and NHCCI-adjusted dollars; base year = earliest year with NHCCI data)."
#| fig-height: 4.5
#| fig-width: 7.5

library(ggplot2)
library(dplyr)
library(tidyr)

plot_df_gf <- gf_final |>
  select(fiscal_year, public_works_nominal, public_works_real) |>
  pivot_longer(
    cols = c(public_works_nominal, public_works_real),
    names_to = "series",
    values_to = "amount"
  ) |>
  mutate(
    series = recode(
      series,
      public_works_nominal = "Nominal ($)",
      public_works_real    = paste0("NHCCI-adjusted (", gf_base_year, " $)")
    )
  )

ggplot(plot_df_gf, aes(x = fiscal_year, y = amount, group = series, linetype = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
  limits = c(100000, NA),
  labels = function(x) paste0("$", format(round(x / 1000, 0), big.mark = ","), "k")
) +
  labs(
    x = "Fiscal year",
    y = "dollars",
    title = paste0(params$town, ": General Fund Public Works spending over time")
  )
```


## Traffic Volume and Congestion

```{r}
#| label: replica-data-prep
#| include: false

library(sf)
library(dplyr)
library(stringr)

### File Paths

# Replica file (town-specific by municipal code)
replica_filename <- paste0("aadt_2024_", params$town_code, ".geojson")
replica_path <- file.path("data_raw", replica_filename)

if (!file.exists(replica_path)) {
  stop(paste0(
    "Replica AADT file mismatch.\n",
    "Expected file: ", replica_filename, "\n",
    "Check that the AADT GeoJSON matches params$town_code."
  ))
}

# Town boundary
bnd_path <- file.path("data_raw", "town_boundaries.geojson")
stopifnot(file.exists(bnd_path))

### Read Data

# Read boundary
bnd <- st_read(bnd_path, quiet = TRUE) |>
  filter(TOWN_ID == params$town_code)

if (nrow(bnd) == 0) {
  stop(paste("No boundary found for town_code =", params$town_code))
}

# Read Replica AADT data
rep <- st_read(replica_path, quiet = TRUE)

### Ensure Projected CRS (required for offsetting geometry)

if (st_is_longlat(rep)) {
  rep <- st_transform(rep, 26986)  # Massachusetts State Plane (meters)
}

if (st_is_longlat(bnd)) {
  bnd <- st_transform(bnd, 26986)
}

### Clip to Town Boundary and filter motorways out

rep_town <- st_intersection(rep, bnd) |>
  filter(highway != "motorway")

if (nrow(rep_town) == 0) {
  stop("No Replica AADT segments intersect town boundary.")
}

### Calculate Truck AADT and Share

rep_town <- rep_town |>
  mutate(
    aadt_single_unit = coalesce(aadt_single_unit, 0),
    aadt_combination = coalesce(aadt_combination, 0),
    truck_aadt = aadt_single_unit + aadt_combination,
    truck_share = ifelse(aadt > 0, truck_aadt / aadt, NA_real_)
  )


```

```{r}
#| label: map-aadt-total
#| fig-cap: "Estimated Annual Average Daily Traffic (AADT), all vehicles (Replica, 2024). Highways (motorway segments) are excluded to emphasize local roadway conditions."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(maptiles)

# Transform to lon/lat for tiles + plotting
bnd_ll <- st_transform(bnd, 4326)

# Filter out motorways (keep motorway_link)
rep_ll <- rep_town |>
  filter(highway != "motorway") |>
  st_transform(4326)

# Manual breaks (edit these as needed)
rep_ll <- rep_ll |>
  mutate(
    aadt_cat = cut(
      aadt,
      breaks = c(-Inf, 500, 1000, 2000, 3000, Inf),
      labels = c("<500", "500 – 1k", "1 – 2k", "2 – 3k", "3k+")
    )
  )

cats <- levels(rep_ll$aadt_cat)

# Line widths (thin -> thick)
lwd_map <- c("<500" = 1, "500 – 1k" = 2, "1 – 2k" = 3, "2 – 3k" = 4, "3k+" = 5)

# Graduated color ramp (light -> dark)
col_map <- c(
  "<500"    = "#F6BDC0",
  "500 – 1k"   = "#F1959B",
  "1 – 2k"  = "#F07470",
  "2 – 3k" = "#EA4C46",
  "3k+"   = "#DC1C13"
)

# Basemap extent from boundary
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.08
ypad <- (bb["ymax"] - bb["ymin"]) * 0.08
bb_exp <- bb
bb_exp["xmin"] <- bb["xmin"] - xpad
bb_exp["xmax"] <- bb["xmax"] + xpad
bb_exp["ymin"] <- bb["ymin"] - ypad
bb_exp["ymax"] <- bb["ymax"] + ypad

tiles <- get_tiles(
  x = bb_exp,
  provider = "Esri.WorldGrayCanvas",
  zoom = 15,
  crop = FALSE
)

plot_tiles(tiles)

# Boundary outline
plot(st_geometry(bnd_ll), add = TRUE, border = "black", lwd = 1)

# Draw segments by category (low -> high, so high draws last)
for (cat in cats) {
  segs <- rep_ll[rep_ll$aadt_cat == cat, ]
  if (nrow(segs) > 0) {
    plot(st_geometry(segs), add = TRUE, col = col_map[[cat]], lwd = lwd_map[[cat]])
  }
}

# Legend (manual position inside map)
bb2 <- st_bbox(bnd_ll)
x_pos <- bb2["xmin"] - .012
y_pos <- bb2["ymin"] + .02

legend(
  x = x_pos, y = y_pos,
  legend = cats,
  col = unname(col_map),
  lwd = unname(lwd_map),
  bty = "o",
  bg = "white",
  box.lwd = 0.5,
  title = "AADT"
)

mtext(get_credit(tiles), side = 1, line = 1, cex = 0.7)
```

```{r}
#| label: map-aadt-truck-share
#| fig-cap: "Truck share of AADT (single-unit + combination trucks as a percentage of total AADT; Replica, 2024). Highways (motorway segments) are excluded to emphasize local roadway conditions."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(maptiles)

# Transform to lon/lat for tiles + plotting
bnd_ll <- st_transform(bnd, 4326)

# Filter out motorways (keep motorway_link) and transform
rep_ll <- rep_town |>
  filter(highway != "motorway") |>
  st_transform(4326)

# Ensure truck share exists (should from prep chunk)
# truck_share is a proportion (0–1). Convert to percent for mapping.
rep_ll <- rep_ll |>
  mutate(
    truck_pct = truck_share * 100,
    truck_cat = cut(
      truck_pct,
      breaks = c(-Inf, 1, 2, 5, 10, Inf),
      labels = c("<1%", "1–2%", "2–5%", "5–10%", "10%+")
    )
  )

cats <- levels(rep_ll$truck_cat)

# Line widths (thin -> thick) to reinforce the categories
lwd_map <- c("<1%" = 1.5, "1–2%" = 2.5, "2–5%" = 3.5, "5–10%" = 4.5, "10%+" = 5.5)

# Color ramp: green (low) -> red (high)
col_map <- c(
  "<1%"    = "#1a9850",
  "1–2%"  = "#91cf60",
  "2–5%" = "#fee08b",
  "5–10%" = "#fc8d59",
  "10%+"   = "#d73027"
)

# Basemap extent from boundary
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.08
ypad <- (bb["ymax"] - bb["ymin"]) * 0.08
bb_exp <- bb
bb_exp["xmin"] <- bb["xmin"] - xpad
bb_exp["xmax"] <- bb["xmax"] + xpad
bb_exp["ymin"] <- bb["ymin"] - ypad
bb_exp["ymax"] <- bb["ymax"] + ypad

tiles <- get_tiles(
  x = bb_exp,
  provider = "Esri.WorldGrayCanvas",
  zoom = 15,
  crop = FALSE
)

plot_tiles(tiles)

# Boundary outline
plot(st_geometry(bnd_ll), add = TRUE, border = "black", lwd = 1)

# Draw segments by category (low -> high so high truck share draws on top)
for (cat in cats) {
  segs <- rep_ll[rep_ll$truck_cat == cat, ]
  if (nrow(segs) > 0) {
    plot(st_geometry(segs), add = TRUE, col = col_map[[cat]], lwd = lwd_map[[cat]])
  }
}

# Legend (manual position inside map; tweak as needed)
bb2 <- st_bbox(bnd_ll)
x_pos <- bb2["xmin"] - .012
y_pos <- bb2["ymin"] + .02

legend(
  x = x_pos, y = y_pos,
  legend = cats,
  col = unname(col_map),
  lwd = unname(lwd_map),
  bty = "o",
  bg = "white",
  box.lwd = 0.5,
  title = "Truck share"
)

mtext(get_credit(tiles), side = 1, line = 1, cex = 0.7)
```

```{r}
#| label: tti-data-prep
#| include: false

library(sf)
library(dplyr)

### --- File Path (town-specific) ---

tti_filename <- paste0("qtr-hourly-speeds_2024_", params$town_code, ".geojson")
tti_path <- file.path("data_raw", tti_filename)

if (!file.exists(tti_path)) {
  stop(paste0(
    "Speed file mismatch.\n",
    "Expected file: ", tti_filename, "\n",
    "Check that the GeoJSON matches params$town_code."
  ))
}

#town boundary
bnd_path <- file.path("data_raw", "town_boundaries.geojson")  # <- change if needed

bnd <- st_read(bnd_path, quiet = TRUE) |>
  filter(TOWN_ID == params$town_code)

### --- Read Speeds Data ---

tti <- st_read(tti_path, quiet = TRUE)

# Ensure projected CRS (required for coordinate shifting)
if (st_is_longlat(tti)) {
  tti <- st_transform(tti, 26986)  # MA State Plane (meters)
}

### --- Identify Peak Period Columns ---

am_cols <- paste0("wkdy_", c(
  "0700","0715","0730","0745",
  "0800","0815","0830","0845"
))

pm_cols <- paste0("wkdy_", c(
  "1600","1615","1630","1645",
  "1700","1715","1730","1745"
))

### --- Compute Peak Speeds ---

tti <- tti |>
  mutate(
    am_peak_speed = rowMeans(across(all_of(am_cols)), na.rm = TRUE),
    pm_peak_speed = rowMeans(across(all_of(pm_cols)), na.rm = TRUE)
  )

### --- Compute Travel Time Index (TTI) ---

tti <- tti |>
  mutate(
    am_tti = free_flow_speed_mph / am_peak_speed,
    pm_tti = free_flow_speed_mph / pm_peak_speed
  )

### --- Offset Directional Links Using Heading Only (no extra sign split) ---

dist_m <- 15  # increase if needed (e.g., 12)

offset_geom_by_heading <- function(geom, heading_deg, dist_m) {
  if (is.na(heading_deg)) heading_deg <- 0

  theta <- heading_deg * pi / 180

  # Direction vector (0=N, 90=E)
  dx <- sin(theta)
  dy <- cos(theta)

  # Perpendicular (right of travel direction)
  ox <-  dy * dist_m
  oy <- -dx * dist_m

  shift_coords <- function(mat) {
    mat[, 1] <- mat[, 1] + ox
    mat[, 2] <- mat[, 2] + oy
    mat
  }

  if (inherits(geom, "LINESTRING")) {
    return(st_linestring(shift_coords(geom)))
  }

  if (inherits(geom, "MULTILINESTRING")) {
    parts <- lapply(geom, shift_coords)
    return(st_multilinestring(parts))
  }

  geom
}

geom_list <- st_geometry(tti)
hdg <- tti$heading

tti_offset <- tti
st_geometry(tti_offset) <- st_sfc(
  lapply(seq_along(geom_list), function(i) {
    offset_geom_by_heading(geom_list[[i]], hdg[i], dist_m)
  }),
  crs = st_crs(tti)
)
```

```{r}
#| label: map-tti-am
#| fig-cap: "AM peak Travel Time Index (TTI), weekdays 7:00–9:00 AM. TTI compares peak travel time to free-flow travel time (TTI = 1.2 indicates ~20% longer travel time than free flow)."
#| fig-height: 8
#| fig-width: 7.5
#| fig-dpi: 300
#| results: "hide"

library(sf)
library(dplyr)
library(maptiles)

# Transform to lon/lat for tiles + plotting
bnd_ll <- st_transform(bnd, 4326)

tti_ll <- tti_offset |>
  # optional: exclude motorways to keep map town-focused (match your AADT maps)
  filter(highway != "motorway") |>
  st_transform(4326)

# Manual TTI bins (edit if needed)
tti_ll <- tti_ll |>
  mutate(
    am_tti_cat = cut(
      am_tti,
      breaks = c(-Inf, 1.05, 1.15, 1.30, 1.50, Inf),
      labels = c("≤1.05", "1.05–1.15", "1.15–1.30", "1.30–1.50", "1.50+")
    )
  )

cats <- levels(tti_ll$am_tti_cat)

# Color ramp: green (low delay) -> red (high delay)
col_map <- c(
  "≤1.05"      = "#1a9850",
  "1.05–1.15"  = "#91cf60",
  "1.15–1.30"  = "#fee08b",
  "1.30–1.50"  = "#fc8d59",
  "1.50+"      = "#d73027"
)

# Basemap extent from boundary
bb <- st_bbox(bnd_ll)
xpad <- (bb["xmax"] - bb["xmin"]) * 0.08
ypad <- (bb["ymax"] - bb["ymin"]) * 0.08
bb_exp <- bb
bb_exp["xmin"] <- bb["xmin"] - xpad
bb_exp["xmax"] <- bb["xmax"] + xpad
bb_exp["ymin"] <- bb["ymin"] - ypad
bb_exp["ymax"] <- bb["ymax"] + ypad

tiles <- get_tiles(
  x = bb_exp,
  provider = "Esri.WorldGrayCanvas",
  zoom = 15,
  crop = FALSE
)

plot_tiles(tiles)

# Boundary outline
plot(st_geometry(bnd_ll), add = TRUE, border = "black", lwd = 1)

# Draw segments by category (low -> high so high congestion draws on top)
for (cat in cats) {
  segs <- tti_ll[tti_ll$am_tti_cat == cat, ]
  if (nrow(segs) > 0) {
    plot(st_geometry(segs), add = TRUE, col = col_map[[cat]], lwd = 1)
  }
}

# Legend (manual position inside map)
bb2 <- st_bbox(bnd_ll)
x_pos <- bb2["xmin"] - .012
y_pos <- bb2["ymin"] + .02

legend(
  x = x_pos, y = y_pos,
  legend = cats,
  col = unname(col_map),
  lwd = 4,
  bty = "o",
  bg = "white",
  box.lwd = 0.5,
  title = "AM Peak TTI"
)

mtext(get_credit(tiles), side = 1, line = 1, cex = 0.7)
```

## Example Goals

# Community Travel Patterns and Accessibility

## Commute and Travel Modes

## Commute Times to Work

## Accessibility

## Walking, Bicycling & Transit Options

## Example Goals

# Safety & Speed Management

## Crash Locations and Severity

## Travel Speed

## Example Goals
